---
title: "Lab 9: Instrumental Variable Analysis"
subtitle: "Causal Machine Learning"
author: "Stijn Vansteelandt and Federico Bertoia, Ghent University"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## 0. Introduction

The goal of this lab is to perform **Instrumental Variable (IV)
analysis** using data from a study of 139 premenopausal women. We are
interested in estimating the causal effect of serum IGFBP-3 levels on
total dense breast tissue area, while accounting for unmeasured
confounding.

Standard regression methods require that all common causes of the
exposure and outcome are measured and adjusted for — an assumption that
is often untestable and likely violated in observational data. When
unmeasured confounders $U$ are present, IV methods can still identify
causal effects, provided a valid instrument $R$ is available.

An **instrumental variable** $R$ must satisfy three assumptions:

1.  **Relevance**: $R$ is associated with the exposure $A$,
2.  **Exclusion restriction**: $R$ affects the outcome $Y$ only through
    $A$,
3.  **Randomization**: $R$ is (conditionally) independent of $Y^a$.

### Estimand

We consider the model $E(Y \mid A, L, R, U) = \psi A + g(L, U)$, where
assumption-lean regression of $Y$ on $R$ given $L$ targets
$$
\beta = \frac{E\{\text{Cov}(R, Y \mid L)\}}{E\{\text{Var}(R \mid L)\}},
$$
and assumption-lean regression of $A$ on $R$ given $L$ targets
$$
\alpha = \frac{E\{\text{Cov}(R, A \mid L)\}}{E\{\text{Var}(R \mid L)\}}.
$$
It follows from the IV assumptions that $\psi = \beta / \alpha$. Since $E\{\text{Var}(R \mid L)\}$
appears in both numerator and denominator, it cancels, giving the **estimand**:
$$
\psi = \frac{E\{\text{Cov}(R, Y \mid L)\}}{E\{\text{Cov}(R, A \mid L)\}}.
$$
This can be estimated via the **Efficient Influence Function (EIF)** estimator:
$$
\hat{\psi} =
\frac{\sum_{i=1}^n \bigl\{R_i - \hat{E}(R_i \mid L_i)\bigr\}
\bigl\{Y_i - \hat{E}(Y_i \mid L_i)\bigr\}}
{\sum_{i=1}^n \bigl\{R_i - \hat{E}(R_i \mid L_i)\bigr\}
\bigl\{A_i - \hat{E}(A_i \mid L_i)\bigr\}}
$$
with standard error estimated as $n^{-1/2}$ times the sample standard
deviation of the efficient influence function.

------------------------------------------------------------------------

### Data

The dataset `igf.dta` contains observations on 139 premenopausal women.

| Variable  | Role           | Description                               |
|-----------|----------------|-------------------------------------------|
| `lucent`  | Outcome $Y$    | Total area of dense tissue in the breast  |
| `igfbp3`  | Exposure $A$   | Serum IGFBP-3 levels                      |
| `igfbp3g` | Instrument $R$ | Putative functional genotype (AA, AC, CC) |
| `age`     | Confounder $L$ | Patient's age                             |
| `afbcat`  | Confounder $L$ | Age at first birth (categorized)          |
| `brfcat`  | Confounder $L$ | Breast feeding duration                   |
| `parcat`  | Confounder $L$ | Number of children                        |
| `ev_pil`  | Confounder $L$ | Ever taken oral contraceptives (binary)   |
| `bmi`     | —              | Body mass index                           |
| `ev_smok` | —              | Ever smoked (binary)                      |

Both the exposure and outcome will be log-transformed throughout to
facilitate interpretation and improve distributional symmetry.

```{r setup-data}
# ── Packages ──────────────────────────────────────────────────────────────────
library(foreign)      
library(dplyr)
library(ggplot2)
library(patchwork)
library(tibble)
library(SuperLearner)
library(lmtest)
library(sandwich)
library(DiagrammeR)
library(AER)

# ── Reproducibility ───────────────────────────────────────────────────────────
set.seed(123)

# ── Load data ─────────────────────────────────────────────────────────────────
data <- read.dta("igf.dta")
N    <- nrow(data)

# ── Derived variables ─────────────────────────────────────────────────────────
data <- data %>%
  mutate(
    log_lucent  = log(lucent),
    log_igfbp3  = log(igfbp3),
    igfbp3g_num = as.numeric(igfbp3g)
  )

glimpse(data)

# ── Analysis variables ────────────────────────────────────────────────────────
Y <- data$log_lucent    # outcome
A <- data$log_igfbp3   # exposure
R <- data$igfbp3g_num  # instrument (numeric: AA=1, AC=2, CC=3)
```

------------------------------------------------------------------------

### Interpreting $\hat\psi$ in the log-log model

Because both outcome and exposure are log-transformed, all results are
reported on the **geometric mean ratio** scale for a 10% increase in
IGFBP-3.

#### Geometric means and log-transforms

The **geometric mean** of a positive random variable $Y$ observed on $n$
units is defined as the $n$-th root of the product of all values:
$$
\text{GM}(Y) = \left(\prod_{i=1}^n y_i\right)^{1/n}.
$$
Taking the log, this becomes an arithmetic mean of logs:
$$
\log \text{GM}(Y) = \frac{1}{n}\sum_{i=1}^n \log y_i = \overline{\log Y}.
$$
In the population, $\overline{\log Y} \to E(\log Y)$ by the law of large
numbers, so exponentiating gives the population geometric mean:
$$
\text{GM}(Y) = \exp\{E(\log Y)\}.
$$
Consequently, a difference on the log scale,
$E(\log Y \mid A) - E(\log Y \mid B)$,
exponentiates to a **ratio of geometric means**:
$$
\exp\{E(\log Y \mid \text{A}) - E(\log Y \mid \text{B})\}
  = \frac{\exp\{E(\log Y \mid \text{A})\}}{\exp\{E(\log Y \mid \text{B})\}}
  = \frac{\text{GM}(Y \mid \text{A})}{\text{GM}(Y \mid \text{B})}.
$$

#### Deriving the geometric mean ratio

Under the partially linear model
$$
E(\log Y \mid A, L) = \psi \log A + g(L),
$$
a 10% increase in $A$ shifts $A \to 1.1A$, so the change in the
conditional mean of $\log Y$ is:
$$
E(\log Y \mid 1.1A, L) - E(\log Y \mid A, L)
  = \psi\log(1.1A) - \psi\log(A)
  = \psi\log\!\left(\frac{1.1A}{A}\right)
  = \psi\log(1.1).
$$
Exponentiating to return to the $Y$ scale, the difference in log-means
becomes a ratio of geometric means:
$$
\frac{\exp\{E(\log Y \mid 1.1A, L)\}}{\exp\{E(\log Y \mid A, L)\}}
  = \exp\!\bigl(\psi\log 1.1\bigr) = 1.1^\psi.
$$
So $1.1^\psi$ is the factor by which the geometric mean of $Y$ is
multiplied when $A$ increases by 10%, irrespective of the baseline value
of $A$ or the covariate profile $L$.

For example, with $\hat\psi = -1.08$:
$$
1.1^{-1.08} = \exp(-1.08 \times \log 1.1) \approx 0.90,
$$
so the geometric mean of $Y$ is multiplied by $0.90$, i.e. reduced by
approximately 10%. 

------------------------------------------------------------------------

## 1. Question 1 — Association of Lucent Area with Genotype

We begin by testing whether the proposed instrument `igfbp3g` is
associated with the outcome `log(lucent)`. Under the IV assumptions, any
such association can only arise through the exposure `igfbp3`. A
significant association is therefore consistent with (but does not
prove) the existence of a causal effect of IGFBP-3 on breast density.

```{r q1-outcome-iv-association}
# ── Linear trend ──────────────────────────────────────────────────────────────
lm_q1 <- lm(log_lucent ~ igfbp3g_num, data = data)
summary(lm_q1)

# ── Factor model (no ordering assumed) ───────────────────────────────────────
lm_q1_cat <- lm(log_lucent ~ factor(igfbp3g), data = data)
summary(lm_q1_cat)

# ── Pearson correlation ───────────────────────────────────────────────────────
cor.test(data$log_lucent, data$igfbp3g_num)

# ── Visualise ────────────────────────────────────────────────────────────────
ggplot(data, aes(x = igfbp3g, y = log_lucent, fill = igfbp3g)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 16, width = 0.45) +
  scale_fill_manual(values = c("tomato", "steelblue", "forestgreen")) +
  theme_minimal() +
  labs(x = "Genotype (igfbp3g)", y = "log(lucent area)",
       title = "Log(Lucent Area) by IGFBP-3 Genotype") +
  theme(legend.position = "none")
```

> **Interpretation:** The genotype is significantly associated with
> log(lucent area) ($p = 0.018$), with each additional C allele
> corresponding to a decrease in log-lucent area of approximately 0.154.
> Under the IV assumptions, this association propagates entirely through
> IGFBP-3 levels — consistent with the instrument having a causal
> downstream effect.

------------------------------------------------------------------------

## 2. Question 2 — Distribution of IGFBP-3 and Relevance of the Instrument

We examine the distribution of the exposure and assess whether the
instrument satisfies the **relevance** assumption — i.e., whether
`igfbp3g` is associated with serum IGFBP-3 after (implicitly) adjusting
for measured confounders. This is the only IV assumption directly
testable from data.

### 2.1 Histogram of IGFBP-3 (raw and log-transformed)

```{r q2-histograms}
p1 <- ggplot(data, aes(x = igfbp3)) +
  geom_histogram(bins = 10, fill = "steelblue", alpha = 0.7, color = "white") +
  theme_minimal() +
  labs(x = "igfbp3", y = "Frequency", title = "Histogram of igfbp3")

p2 <- ggplot(data, aes(x = log_igfbp3)) +
  geom_histogram(bins = 10, fill = "steelblue", alpha = 0.7, color = "white") +
  theme_minimal() +
  labs(x = "log_igfbp3", y = "Frequency", title = "Histogram of log_igfbp3")

p1 + p2
```

> The raw distribution of IGFBP-3 is right-skewed; the log
> transformation yields an approximately symmetric distribution, making
> it more suitable for linear models.

### 2.2 Distribution of log(IGFBP-3) by genotype

```{r q2-by-genotype}
# ── Summary statistics by genotype ────────────────────────────────────────────
data %>%
  group_by(igfbp3g) %>%
  summarise(
    n    = n(),
    mean = round(mean(log_igfbp3), 3),
    sd   = round(sd(log_igfbp3),   3),
    min  = round(min(log_igfbp3),  3),
    max  = round(max(log_igfbp3),  3)
  )

# ── Overlapping histograms ────────────────────────────────────────────────────
ggplot(data, aes(x = log_igfbp3, fill = igfbp3g, colour = igfbp3g)) +
  geom_histogram(alpha = 0.4, position = "identity", bins = 10) +
  scale_fill_manual(values   = c("tomato", "steelblue", "forestgreen")) +
  scale_colour_manual(values = c("tomato", "steelblue", "forestgreen")) +
  theme_minimal() +
  labs(x = "log_igfbp3", y = "Count",
       title = "Distribution of log(IGFBP-3) by genotype",
       fill = "igfbp3g", colour = "igfbp3g")
```

### 2.3 Testing relevance

```{r q2-relevance}
# ── Linear regression of log(igfbp3) on genotype ─────────────────────────────
lm_q2 <- lm(log_igfbp3 ~ igfbp3g_num, data = data)
summary(lm_q2)

# ── As factor, with F-test ────────────────────────────────────────────────────
lm_q2_cat <- lm(log_igfbp3 ~ factor(igfbp3g), data = data)
anova(lm_q2_cat)
```

> **Interpretation:** The genotype is significantly associated with
> log(IGFBP-3) ($p = 0.009$), supporting the relevance assumption. Each
> additional C allele is associated with a decrease of approximately
> 0.063 in log(IGFBP-3). This is the only IV assumption directly
> testable from data; the exclusion restriction and randomization
> assumption require subject-matter justification.

------------------------------------------------------------------------

## 3. Question 3 — DAG and IV Assumptions

```{r q3-dag}
grViz("
digraph IV_DAG {
  graph [layout = neato, bgcolor = transparent, splines = curved]
  node [shape = plaintext, fontname = 'Helvetica', fontsize = 16]
  R [pos = '0,0!',  label = 'R\n(igfbp3g)']
  A [pos = '3,0!',  label = 'A\n(log igfbp3)']
  Y [pos = '6,0!',  label = 'Y\n(log lucent)']
  U [pos = '4.5,2!', label = 'U']
  L [pos = '3,-2!', label = 'L\n(age, afbcat,\nbrfcat, parcat, ev_pil)']

  edge [arrowhead = vee, arrowsize = 0.8, penwidth = 1.6, color = '#444444']
  R -> A
  A -> Y
  U -> A
  U -> Y
  L -> A
  L -> Y

  edge [arrowhead = vee, arrowsize = 0.7, penwidth = 1.2,
        style = dashed, color = '#888888']
  R -> Y [style = dashed, constraint = false]
  R -> L [style = dashed]
}
")
```

The three IV assumptions in this setting are:

1.  **Relevance** $R \not\!\perp\!\!\!\perp A (\mid L)$: The IGFBP3g
    genotype must be associated with serum IGFBP-3 levels (after
    adjusting for observed covariates $L$). This was verified empirically
    in Question 2.

2.  **Exclusion restriction** $R \perp\!\!\!\perp Y \mid A, (L), U$: The
    genotype should not affect dense breast tissue area through any
    pathway other than serum IGFBP-3. This assumption is *potentially
    violated* if the genotype affects fertility-related traits (such as
    parity or age at first birth), which in turn influence breast tissue
    density.

3.  **Randomization** $R \perp\!\!\!\perp U (\mid L)$: The genotype is
    randomized by Mendelian inheritance — transmission of alleles from
    parent to offspring is random, so it should not share common causes
    with the unmeasured confounders $U$ (after conditioning on $L$). This
    assumption is not testable from data and relies on biological
    reasoning.

------------------------------------------------------------------------

## 4. Question 4 — IV Estimate via Efficient Influence Function (no covariates)

We estimate $\psi$ using the EIF estimator without adjusting for any
covariates $L$. In this case $\hat{E}(R \mid L) = \bar{R}$,
$\hat{E}(A \mid L) = \bar{A}$, and $\hat{E}(Y \mid L) = \bar{Y}$, so the
estimator reduces to:

$$
\hat{\psi} =
\frac{\sum_i (R_i - \bar{R})(Y_i - \bar{Y})}
     {\sum_i (R_i - \bar{R})(A_i - \bar{A})}
$$

```{r q4-iv-no-covariates}
# ── Point estimate ────────────────────────────────────────────────────────────
beta_q4 <- mean((R - mean(R)) * (Y - mean(Y))) /
           mean((R - mean(R)) * (A - mean(A)))

# ── Standard error via influence function ─────────────────────────────────────
EIC_beta_q4 <- (R - mean(R)) * (Y - mean(Y) - beta_q4 * (A - mean(A))) /
               mean((R - mean(R)) * (A - mean(A)))
se_beta_q4  <- sd(EIC_beta_q4) / sqrt(N)

# ── 95% confidence interval ───────────────────────────────────────────────────
beta_q4_lb <- beta_q4 - 1.96 * se_beta_q4
beta_q4_ub <- beta_q4 + 1.96 * se_beta_q4

# ── Results on geometric mean ratio scale (10% increase in A) ─────────────────
results_q4 <- tibble(
  Method   = "EIF (no covariates)",
  Estimate = round(exp(beta_q4    * log(1.1)), 3),
  CI_Lower = round(exp(beta_q4_lb * log(1.1)), 3),
  CI_Upper = round(exp(beta_q4_ub * log(1.1)), 3)
)

# ── 2SLS comparison (no covariates) ──────────────────────────────────────────
iv_q4    <- ivreg(log_lucent ~ log_igfbp3 | igfbp3g_num, data = data)
iv_q4_ct <- coeftest(iv_q4, vcov = sandwich)
iv_q4_ci <- coefci(iv_q4,   vcov = sandwich)["log_igfbp3", ]

beta_q4_2sls    <- iv_q4_ct["log_igfbp3", 1]
beta_q4_2sls_lb <- iv_q4_ci[1]
beta_q4_2sls_ub <- iv_q4_ci[2]

results_q4 <- rbind(results_q4, tibble(
  Method   = "2SLS (no covariates)",
  Estimate = round(exp(beta_q4_2sls    * log(1.1)), 3),
  CI_Lower = round(exp(beta_q4_2sls_lb * log(1.1)), 3),
  CI_Upper = round(exp(beta_q4_2sls_ub * log(1.1)), 3)
))

knitr::kable(results_q4, digits = 3,
             caption = "Q4 — EIF vs 2SLS (no covariates): GMR for 10% increase in IGFBP-3")
```

> **Interpretation:** A 10% increase in serum IGFBP-3 multiplies the
> geometric mean of dense breast tissue area by approximately $1.1^{\hat\psi}
> \approx 1.261$, i.e. an increase of roughly 26.1% [95% CI: 0.1% to
> 58.9%]. The confidence interval is wide, reflecting the limited sample
> size and weak instrument strength.

------------------------------------------------------------------------

## 5. Question 5 — DML IV Estimate with Covariates (SuperLearner)

A concern with the unadjusted IV estimate is that `parcat` and `afbcat`
(parity and age at first birth) may lie on the causal pathway from
genotype to outcome — violating the exclusion restriction. We address
this by conditioning on $L = (\texttt{afbcat}, \texttt{parcat})$,
thereby blocking that pathway.

We use **Debiased Machine Learning** with SuperLearner to estimate the
three nuisance functions $\hat{E}(Y \mid L)$, $\hat{E}(A \mid L)$ and
$\hat{E}(R \mid L)$. Given the small sample ($n = 139$), cross-fitting
is omitted and a restricted library is used.

```{r q5-dml-iv}
# ── Confounder set blocking the problematic pathway ───────────────────────────
L_5 <- data %>% dplyr::select(afbcat, parcat)

SL.library <- c("SL.glm", "SL.ranger", "SL.earth")

# ── Nuisance models ───────────────────────────────────────────────────────────
sl_Y_q5 <- SuperLearner(Y = Y, X = L_5, SL.library = SL.library)
sl_A_q5 <- SuperLearner(Y = A, X = L_5, SL.library = SL.library)
sl_R_q5 <- SuperLearner(Y = R, X = L_5, SL.library = SL.library)

# ── Nuisance predictions ──────────────────────────────────────────────────────
m_Y_q5 <- predict(sl_Y_q5, newdata = L_5, onlySL = TRUE)$pred
m_A_q5 <- predict(sl_A_q5, newdata = L_5, onlySL = TRUE)$pred
m_R_q5 <- predict(sl_R_q5, newdata = L_5, onlySL = TRUE)$pred

# ── EIF estimator with covariates ─────────────────────────────────────────────
beta_q5 <- mean((R - m_R_q5) * (Y - m_Y_q5)) /
           mean((R - m_R_q5) * (A - m_A_q5))

# ── Standard error ────────────────────────────────────────────────────────────
EIC_beta_q5 <- (R - m_R_q5) * (Y - m_Y_q5 - beta_q5 * (A - m_A_q5)) /
               mean((R - m_R_q5) * (A - m_A_q5))
se_beta_q5  <- sd(EIC_beta_q5) / sqrt(N)

# ── 95% confidence interval ───────────────────────────────────────────────────
beta_q5_lb <- beta_q5 - 1.96 * se_beta_q5
beta_q5_ub <- beta_q5 + 1.96 * se_beta_q5

# ── Results ───────────────────────────────────────────────────────────────────
results_q5 <- tibble(
  Method   = "EIF (afbcat + parcat)",
  Estimate = round(exp(beta_q5    * log(1.1)), 3),
  CI_Lower = round(exp(beta_q5_lb * log(1.1)), 3),
  CI_Upper = round(exp(beta_q5_ub * log(1.1)), 3)
)

# ── 2SLS comparison (afbcat + parcat) ────────────────────────────────────────
iv_q5    <- ivreg(log_lucent ~ log_igfbp3 + afbcat + parcat |
                    igfbp3g_num + afbcat + parcat, data = data)
iv_q5_ct <- coeftest(iv_q5, vcov = sandwich)
iv_q5_ci <- coefci(iv_q5,   vcov = sandwich)["log_igfbp3", ]

beta_q5_2sls    <- iv_q5_ct["log_igfbp3", 1]
beta_q5_2sls_lb <- iv_q5_ci[1]
beta_q5_2sls_ub <- iv_q5_ci[2]

results_q5 <- rbind(results_q5, tibble(
  Method   = "2SLS (afbcat + parcat)",
  Estimate = round(exp(beta_q5_2sls    * log(1.1)), 3),
  CI_Lower = round(exp(beta_q5_2sls_lb * log(1.1)), 3),
  CI_Upper = round(exp(beta_q5_2sls_ub * log(1.1)), 3)
))

knitr::kable(results_q5, digits = 3,
             caption = "Q5 — EIF vs 2SLS (afbcat + parcat): GMR for 10% increase in IGFBP-3")
```

> **Interpretation:** After conditioning on `afbcat` and `parcat` to
> block the potentially problematic pathway from genotype to outcome, a
> 10% increase in serum IGFBP-3 multiplies the geometric mean of dense
> breast tissue area by approximately $1.1^{\hat\psi} \approx 1.268$,
> i.e. an increase of roughly 26.8% [95% CI: 1.4% to 58.5%]. The
> estimate is close to Question 4 and the
> interval no longer includes zero.

------------------------------------------------------------------------

## 6. Question 6 — Assumption-Lean Regression (Comparison)

For comparison, we estimate the **association** (not necessarily causal
effect) between log(IGFBP-3) and log(lucent area) using
**assumption-lean linear regression** — adjusting for the full
confounder set $L = (\texttt{age}, \texttt{afbcat}, \texttt{brfcat},
\texttt{parcat}, \texttt{ev\_pil})$. 
```{r q6-assumption-lean}
# ── Full confounder set ───────────────────────────────────────────────────────
L_6 <- data %>% dplyr::select(age, afbcat, brfcat, parcat, ev_pil)

# ── Nuisance models ───────────────────────────────────────────────────────────
sl_Y_q6 <- SuperLearner(Y = Y, X = L_6, SL.library = SL.library)
sl_A_q6 <- SuperLearner(Y = A, X = L_6, SL.library = SL.library)

m_Y_q6 <- predict(sl_Y_q6, newdata = L_6, onlySL = TRUE)$pred
m_A_q6 <- predict(sl_A_q6, newdata = L_6, onlySL = TRUE)$pred

# ── EIF estimator ──────────────────────────────────────────────────
A_res <- A - m_A_q6
Y_res <- Y - m_Y_q6

beta_q6 <- mean(A_res * Y_res) / mean(A_res^2)

# ── Standard error via influence function ─────────────────────────────────────
EIC_beta_q6 <- A_res * (Y_res - beta_q6 * A_res) / mean(A_res^2)
se_beta_q6  <- sd(EIC_beta_q6) / sqrt(N)

# ── 95% CI ────────────────────────────────────────────────────────────────────
beta_q6_lb <- beta_q6 - 1.96 * se_beta_q6
beta_q6_ub <- beta_q6 + 1.96 * se_beta_q6

# ── Results ───────────────────────────────────────────────────────────────────
results_q6 <- tibble(
  Method   = "Assumption-lean regression (full L)",
  Estimate = round(exp(beta_q6    * log(1.1)), 3),
  CI_Lower = round(exp(beta_q6_lb * log(1.1)), 3),
  CI_Upper = round(exp(beta_q6_ub * log(1.1)), 3)
)

knitr::kable(results_q6, digits = 3,
             caption = "Q6 — Assumption-lean regression: GMR for 10% increase in IGFBP-3")
```

> **Interpretation:** The assumption-lean regression yields a much
> smaller association: a 10% increase in IGFBP-3 multiplies the
> geometric mean of dense breast tissue area by approximately
> $1.1^{\hat\psi} \approx 1.046$, i.e. an increase of roughly 4.6%
> [95% CI: 0.1% to 9.4%], which is not statistically significant. This
> contrasts sharply with the IV estimates ($\approx 25\%$), suggesting
> that unmeasured confounding substantially attenuates the observational
> association.

------------------------------------------------------------------------

## 7. Summary of Results

```{r summary-table}
results_all <- tibble(
  Method = c(
    "IV — EIF (no covariates)",
    "IV — 2SLS (no covariates)",
    "IV — EIF (afbcat + parcat)",
    "IV — 2SLS (afbcat + parcat)",
    "Assumption-lean regression (full L)"
  ),
  Type = c("IV", "IV", "IV", "IV", "Regression"),
  Estimate = exp(c(beta_q4, beta_q4_2sls, beta_q5, beta_q5_2sls, beta_q6) * log(1.1)),
  Lower    = exp(c(beta_q4_lb, beta_q4_2sls_lb, beta_q5_lb, beta_q5_2sls_lb, beta_q6_lb) * log(1.1)),
  Upper    = exp(c(beta_q4_ub, beta_q4_2sls_ub, beta_q5_ub, beta_q5_2sls_ub, beta_q6_ub) * log(1.1))
)

knitr::kable(results_all, digits = 3,
             caption = "All estimates: GMR for a 10% increase in IGFBP-3")
```

```{r forest-plot}
results_all %>%
  mutate(Method = factor(Method, levels = rev(Method))) %>%
  ggplot(aes(x = Estimate, y = Method, colour = Type)) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey60") +
  geom_errorbarh(aes(xmin = Lower, xmax = Upper), height = 0.25, linewidth = 0.8) +
  geom_point(size = 3) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1, scale = 100)) +
  scale_colour_manual(values = c("IV" = "tomato", "Regression" = "steelblue")) +
  theme_minimal(base_size = 13) +
  labs(
    x      = "Geometric mean ratio (10% increase in IGFBP-3)",
    y      = NULL,
    colour = NULL,
    title  = "IV vs Regression Estimates — IGFBP-3 Effect on Breast Density"
  ) +
  theme(legend.position = "bottom")
```

> **Overall conclusion:** The IV estimates suggest a substantial
> positive causal effect of serum IGFBP-3 on dense breast tissue area.
> A 10% increase in IGFBP-3 multiplies the geometric mean lucent area
> by roughly $1.1^\psi \approx 1.26$, i.e. an increase of $\sim 26\%$.
> The assumption-lean regression estimate is substantially smaller, pointing to negative unmeasured confounding that
> masks the causal effect in a standard observational analysis. 