---
title: "Lab 5: Average Treatment Effect"
subtitle: "Causal Machine Learning"
author: "Stijn Vansteelandt and Federico Bertoia, Ghent University"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## 0. Introduction

The goal of this lab is to estimate the **Average Treatment Effect
(ATE)**, defined as:

$$
E[Y^1 - Y^0],
$$

where $Y^a$ denotes the potential outcome had treatment been set to
$A = a$. Under the standard causal assumptions of **consistency**,
**positivity**, and **conditional exchangeability**, the ATE is
nonparametrically identified from observed data as:

$$
\text{ATE} = E\Big[E[Y \mid A=1, X] - E[Y \mid A=0, X]\Big].
$$

We implement and compare four estimators:

1.  **G-Computation**
2.  **IPW** (Inverse Probability Weighting)
3.  **AIPW** (Augmented IPW)
4.  **TMLE** (Targeted Maximum Likelihood Estimation)

------------------------------------------------------------------------

### Data

The dataset comes from an observational study conducted at Ohio Heart
Health (1997), comprising **996 patients** who underwent Percutaneous
Coronary Intervention (PCI). Patients received PCI either with or
without adjunct therapy using **abciximab**, a glycoprotein IIb/IIIa
inhibitor that reduces platelet aggregation.

Importantly, abciximab was preferentially administered to **more
severely diseased patients**, introducing *confounding* — a classic
challenge in observational studies. All patients were followed for at
least 6 months by the Lindner Center staff, and the binary outcome of
interest is **survival at 6 months (0 if survived, 1 otherwise)**.

```{r setup-data}
# ── Packages ──────────────────────────────────────────────────────────────────
library(twang)
library(tibble)
library(dplyr)
library(tidyr) 
library(ggplot2)
library(SuperLearner)
library(tmle)

# ── Reproducibility ───────────────────────────────────────────────────────────
set.seed(123)

# ── Load data ─────────────────────────────────────────────────────────────────
data(lindner)

lindner <- lindner %>%
  mutate(survival_binary = ifelse(sixMonthSurvive == FALSE, 1, 0)) %>%
  select(-sixMonthSurvive, -lifepres)

# ── Initial inspection ────────────────────────────────────────────────────────
glimpse(lindner)

# Confirm no missing values
stopifnot(all(colSums(is.na(lindner)) == 0))
```

## 1. Exploratory Data Analysis

Before estimating the ATE, we briefly explore the data to understand the
distribution of the treatment, outcome, and key covariates.

### 1.1 Treatment and outcome distribution

```{r marginal distributions treatment and outcome}
library(patchwork)

p1 <- lindner %>%
  mutate(survival_binary = factor(survival_binary, labels = c("Survived", "Died"))) %>%
  ggplot(aes(x = survival_binary, fill = survival_binary)) +
  geom_bar(alpha = 0.8, width = 0.4) +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  theme_minimal() +
  labs(x = NULL, y = "Count", fill = "Outcome", title = "Outcome")

p2 <- lindner %>%
  mutate(abcix = factor(abcix, labels = c("No abciximab", "Abciximab"))) %>%
  ggplot(aes(x = abcix, fill = abcix)) +
  geom_bar(alpha = 0.8, width = 0.4) +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  theme_minimal() +
  labs(x = NULL, y = "Count", fill = "Treatment", title = "Treatment")

p1 + p2
```

```{r eda-treatment-outcome}
lindner %>%
  mutate(
    abcix           = factor(abcix, labels = c("No abciximab", "Abciximab")),
    survival_binary = factor(survival_binary, labels = c("Survived", "Died"))
  ) %>%
  count(abcix, survival_binary) %>%
  group_by(abcix) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = abcix, y = prop, fill = survival_binary)) +
  geom_col(position = "dodge", alpha = 0.8, width = 0.5) +
  geom_text(aes(label = scales::percent(prop, accuracy = 0.1)),
            position = position_dodge(width = 0.5), vjust = -0.5, size = 3.5) +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1.05)) +
  theme_minimal() +
  labs(x = NULL, y = "Proportion", fill = "Outcome",
       title = "Survival at 6 months by treatment group")
```

### 1.2 Covariate distributions 

```{r eda-covariates}

cat_vars <- c("stent", "female", "diabetic", "acutemi", "ves1proc")

lindner %>%
  select(-survival_binary) %>%
  mutate(abcix = factor(abcix, labels = c("No abciximab", "Abciximab"))) %>%
  pivot_longer(cols = -abcix, names_to = "variable", values_to = "value") %>%
  mutate(is_cat = variable %in% cat_vars) %>%
  ggplot(aes(x = value, fill = abcix, color = abcix)) +
  geom_bar(
    data  = ~ filter(.x, is_cat),
    fill  = "steelblue", color = "steelblue", alpha = 0.6,
    width = 0.4
  ) +
  geom_density(
    data  = ~ filter(.x, !is_cat),
    fill  = "steelblue", color = "steelblue", alpha = 0.3, linewidth = 0.8
  ) +
  facet_wrap(~variable, scales = "free") +
  theme_minimal() +
  labs(x = NULL, y = "Count / Density")
```

## 2. Unadjusted Risk Difference and Odds Ratio

Before turning to "causal" estimators, we compute two simple unadjusted
measures of association between treatment and outcome: the **risk
difference** (RD) and the **odds ratio** (OR).

### 2.1 Risk Difference

The risk difference is the difference between the risk of an outcome in
the treated group and non treated group.

$$
E[Y \mid A=1] - E[Y \mid A=0] = P(Y=1 \mid A=1) - P(Y=1 \mid A=0)
$$

It can be estimated via a linear probability model:

$$ E(Y \mid A) = \alpha + \beta A $$

where $\hat{\beta}$ directly estimates the risk difference.

```{r risk difference}

# ── Model formulas ────────────────────────────────────────────────────────────
unadjusted <- survival_binary ~ abcix

# ── Risk Difference (linear probability model) ────────────────────────────────
lpm_fit <- glm(unadjusted, data = lindner, family = gaussian())
summary(lpm_fit)
confint(lpm_fit)

rd    <- coef(lpm_fit)["abcix"]
ci_rd <- confint(lpm_fit)["abcix", ]

cat("Risk Difference:", round(rd, 4), "\n")
cat("95% CI: [", round(ci_rd[1], 4), ",", round(ci_rd[2], 4), "]\n")

```

### 2.2 Odds Ratio

The odds ratio is estimated via logistic regression:

$$
\text{logit}\, P(Y = 1 \mid A) = \log \frac{P(Y=1 \mid A)}{P(Y=0 \mid A)} = \alpha^* + \beta^* A
$$

The odds ratio is then recovered as:

$$
\widehat{\text{OR}} = \frac{\color{green}{\text{odds}(Y=1 \mid A=1)}}{\color{blue}{\text{odds}(Y=1 \mid A=0)}} = \color{green}{\frac{P(Y=1 \mid A=1)}{P(Y=0 \mid A=1)}} \times \color{blue}{\frac{P(Y=0 \mid A=0)}{P(Y=1 \mid A=0)}}
= \frac{\color{green}{\exp(\hat{\alpha}^* + \hat{\beta}^*)}}{\color{blue}{\exp(\hat{\alpha}^*)}} = \exp(\hat{\beta}^*)
$$

```{r odds ratio}
# ── Odds Ratio (logistic regression) ─────────────────────────────────────────
logit_fit <- glm(unadjusted, data = lindner, family = binomial())
summary(logit_fit)

or    <- exp(coef(logit_fit)["abcix"])
ci_or <- exp(confint(logit_fit)["abcix", ])

cat("Odds Ratio:", round(or, 4), "\n")
cat("95% CI: [", round(ci_or[1], 4), ",", round(ci_or[2], 4), "]\n")
```

## 3. Setup

We define the outcome $Y$, treatment $A$, and covariate matrix $X$ used
by all four estimators.

```{r setup-estimators}
# ── Outcome, treatment, covariates ────────────────────────────────────────────
Y <- lindner$survival_binary
A <- lindner$abcix
X <- lindner %>% select(-survival_binary, -abcix)

N <- nrow(lindner)

# ── Candidate learners ────────────────────────────────────────────────────────
SL.library <- c("SL.mean", "SL.glm", "SL.ranger", "SL.gam", "SL.earth")

# ── Cross-fitting folds ───────────────────────────────────────────────────────
n_folds <- 5
fold_id <- sample(rep(1:n_folds, length.out = N))
folds   <- split(seq_len(N), fold_id)
```

## 4. G-Computation

G-Computation proceeds in two steps. First, fit an outcome model
$\hat{Q}(a, x) = \hat{E}[Y \mid A = a, X = x]$. Second, plug in $A = 1$
and $A = 0$ for every unit and average the difference in predicted
potential outcomes:

$$
\widehat{\text{ATE}}_{\text{G-comp}} = \frac{1}{n} \sum_{i=1}^{n} 
\left[ \hat{Q}(1, X_i) - \hat{Q}(0, X_i) \right].
$$

### 4.1 Parametric outcome model

We first fit a parametric logistic regression for $Q(a, x)$ and obtain
the ATE by hand. Standard errors are then computed via `stdReg`, which
uses a sandwich estimator for the variance of $\hat{E}[Y^1]$ and
$\hat{E}[Y^0]$:

$$
\widehat{\text{var}}\left(\hat{E}[Y^1 - Y^0]\right) = 
\widehat{\text{var}}\left(\hat{E}[Y^1]\right) + 
\widehat{\text{var}}\left(\hat{E}[Y^0]\right) - 
2\,\widehat{\text{cov}}\left(\hat{E}[Y^1], \hat{E}[Y^0]\right)
$$

```{r g-computation-parametric}
# ── Model formula ─────────────────────────────────────────────────────────────
formula_gcomp <- as.formula(paste("Y ~ A +", paste(colnames(X), collapse = " + ")))

# ── Fit parametric outcome model ──────────────────────────────────────────────
glm_Q <- glm(formula_gcomp, family = binomial(), data = data.frame(Y, A, X))

# ── Predict potential outcomes by hand ───────────────────────────────────────
dat0 <- data.frame(Y, A = 0, X)
dat1 <- data.frame(Y, A = 1, X)

Q0_par <- predict(glm_Q, newdata = dat0, type = "response")
Q1_par <- predict(glm_Q, newdata = dat1, type = "response")

ate_gcomp_par <- mean(Q1_par - Q0_par)
cat("G-Computation ATE (parametric, by hand):", round(ate_gcomp_par, 4), "\n")

# ── Standard error via stdReg (sandwich estimator) ────────────────────────────
library(stdReg)
fit_std <- stdGlm(fit = glm_Q, data = data.frame(Y, A, X), X = "A")
print(summary(fit_std))

ate_gcomp_par_stdreg <- fit_std$est[2] - fit_std$est[1];
ate_gcomp_par_se <- sqrt(fit_std$vcov[1] + fit_std$vcov[4] - 2 * fit_std$vcov[3])
ci_gcomp_par    <- ate_gcomp_par + c(-1, 1) * 1.96 * ate_gcomp_par_se

cat("G-Computation ATE (parametric, by stdReg):", round(ate_gcomp_par_stdreg, 4), "\n")
cat("SE:", round(ate_gcomp_par_se, 4), "\n")
cat("95% CI: [", round(ci_gcomp_par[1], 4), ",", round(ci_gcomp_par[2], 4), "]\n")
```

### 4.2 Non-parametric outcome model (Super Learner)

We now replace the parametric logistic regression with a Super Learner
ensemble, which avoids model misspecification but introduces **plug-in
bias** — nonparametric estimators converge at slower rates, making valid
confidence intervals unavailable in this setting.

```{r g-computation}
# ── Without cross-fitting ─────────────────────────────────────────────────────
sl_Q <- SuperLearner(
  Y          = Y,
  X          = data.frame(A = A, X),
  SL.library = SL.library,
  family     = binomial(),
  method     = "method.NNLS",
  cvControl  = list(V = 5)
)

Q1 <- predict(sl_Q, newdata = data.frame(A = 1, X))$pred
Q0 <- predict(sl_Q, newdata = data.frame(A = 0, X))$pred

ate_gcomp <- mean(Q1 - Q0)
cat("G-Computation ATE (no cross-fitting):", round(ate_gcomp, 4), "\n")

# ── With cross-fitting ────────────────────────────────────────────────────────
Q1_cf <- numeric(N)
Q0_cf <- numeric(N)

for (k in 1:n_folds) {

  test  <- folds[[k]]
  train <- setdiff(seq_len(N), test)

  sl_Q_cf <- SuperLearner(
    Y          = Y[train],
    X          = data.frame(A = A[train], X[train, ]),
    SL.library = SL.library,
    family     = binomial(),
    method     = "method.NNLS",
    cvControl  = list(V = 5)
  )

  Q1_cf[test] <- predict(sl_Q_cf, newdata = data.frame(A = 1, X[test, ]))$pred
  Q0_cf[test] <- predict(sl_Q_cf, newdata = data.frame(A = 0, X[test, ]))$pred
}

ate_gcomp_cf <- mean(Q1_cf - Q0_cf)
cat("G-Computation ATE (cross-fitting):", round(ate_gcomp_cf, 4), "\n")
```

## 5. IPW

The Inverse Probability Weighting (IPW) estimator reweights observations
by the inverse of their propensity score
$\hat{\pi}(x) = \hat{P}(A = 1 \mid X = x)$, upweighting units whose
treatment assignment was unlikely given their covariates.

$$
\widehat{\text{ATE}}_{\text{IPW}} = \frac{1}{n} \sum_{i=1}^{n}
\left( \frac{A_i Y_i}{\hat{\pi}(X_i)} - \frac{(1-A_i) Y_i}{1 - \hat{\pi}(X_i)} \right)
$$

An alternative is the **Hájek (self-normalized) IPW estimator**, which
normalizes the weights within each treatment arm:

$$
\widehat{\text{ATE}}_{\text{Hájek}} = \hat{E}[Y^1] - \hat{E}[Y^0] = \frac{\sum_{i=1}^{n} \frac{A_i}{\hat{\pi}(X_i)} Y_i}{\sum_{i=1}^{n} \frac{A_i}{\hat{\pi}(X_i)}} - \frac{\sum_{i=1}^{n} \frac{1-A_i}{1-\hat{\pi}(X_i)} Y_i}{\sum_{i=1}^{n} \frac{1-A_i}{1-\hat{\pi}(X_i)}}
$$

This is equivalent to a weighted linear regression of $Y$ on $A$ with
stabilized weights
$w_i = A_i/\hat{\pi}(X_i) + (1-A_i)/(1-\hat{\pi}(X_i))$, since `lm()`
internally normalizes weights within each group. The Hájek estimator is
generally preferred in finite samples as it is less sensitive to extreme
propensity scores.

### 5.1 Parametric propensity score model

We first estimate $\hat{\pi}(x)$ using a parametric logistic regression.
The ATE is obtained via a weighted linear regression of $Y$ on $A$, with
stabilized IPW weights
$w_i = A_i / \hat{\pi}(X_i) + (1 - A_i) / (1 - \hat{\pi}(X_i))$.
Standard errors are computed using the heteroskedasticity-consistent HC
sandwich estimator.

```{r ipw-parametric}
# ── Fit parametric propensity score model ─────────────────────────────────────
library(sandwich)

formula_pi  <- as.formula(paste("A ~", paste(colnames(X), collapse = " + ")))
glm_pi     <- glm(formula_pi, family = binomial(), data = data.frame(A, X))
pi_hat_par <- glm_pi$fitted.values

# ── Positivity check ──────────────────────────────────────────────────────────
# ── Positivity check ──────────────────────────────────────────────────────────
summary(pi_hat_par)

data.frame(pi_hat = pi_hat_par, A = factor(A, labels = c("No abciximab", "Abciximab"))) %>%
  ggplot(aes(x = pi_hat, fill = A, color = A)) +
  geom_density(alpha = 0.3, linewidth = 0.8) +
  scale_fill_manual(values  = c("steelblue", "tomato")) +
  scale_color_manual(values = c("steelblue", "tomato")) +
  theme_minimal() +
  labs(x     = expression(hat(pi)(X)),
       y     = "Density",
       fill  = "Treatment",
       color = "Treatment",
       title = "Propensity scores by treatment group (parametric)")

data.frame(
  weight = c(1 / pi_hat_par[A == 1], 1 / (1 - pi_hat_par[A == 0])),
  A      = factor(c(rep(1, sum(A == 1)), rep(0, sum(A == 0))),
                  labels = c("No abciximab", "Abciximab"))
) %>%
  ggplot(aes(x = A, y = weight, fill = A)) +
  geom_boxplot(alpha = 0.6, outlier.shape = 21) +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  theme_minimal() +
  labs(x = NULL, y = "Inverse weight",
       fill  = "Treatment",
       title = "Distribution of inverse weights by treatment group (parametric)")

# ── Weighted linear regression ────────────────────────────────────────────────
weights_par <- A / pi_hat_par + (1 - A) / (1 - pi_hat_par)
fit_ipw_par <- lm(Y ~ A, weights = weights_par)

ate_ipw_par    <- coef(fit_ipw_par)["A"]
ate_ipw_par_se <- sqrt(vcovHC(fit_ipw_par, type = "HC")["A", "A"])
ci_ipw_par     <- ate_ipw_par + c(-1, 1) * 1.96 * ate_ipw_par_se

cat("IPW ATE (parametric, weighted regression):", round(ate_ipw_par, 4), "\n")
cat("SE:", round(ate_ipw_par_se, 4), "\n")
cat("95% CI: [", round(ci_ipw_par[1], 4), ",", round(ci_ipw_par[2], 4), "]\n")
```

### 5.2 Non-parametric propensity score model (Super Learner)

We now replace the parametric logistic regression with a Super Learner
ensemble. For each estimator we report both the Horvitz-Thompson
estimator (direct plug-in) and the weighted linear regression, which
provides standard errors via the HC sandwich estimator.

```{r ipw}
# ── Without cross-fitting ─────────────────────────────────────────────────────
sl_pi <- SuperLearner(
  Y          = A,
  X          = X,
  SL.library = SL.library,
  family     = binomial(),
  method     = "method.NNLS",
  cvControl  = list(V = 5)
)

pi_hat <- predict(sl_pi, newdata = X)$pred

# ── Positivity check ──────────────────────────────────────────────────────────
summary(pi_hat)

data.frame(pi_hat = pi_hat, A = factor(A, labels = c("No abciximab", "Abciximab"))) %>%
  ggplot(aes(x = pi_hat, fill = A, color = A)) +
  geom_density(alpha = 0.3, linewidth = 0.8) +
  scale_fill_manual(values  = c("steelblue", "tomato")) +
  scale_color_manual(values = c("steelblue", "tomato")) +
  theme_minimal() +
  labs(x     = expression(hat(pi)(X)),
       y     = "Density",
       fill  = "Treatment",
       color = "Treatment",
       title = "Propensity scores by treatment group (no cross-fitting)")

data.frame(
  weight = c(1 / pi_hat[A == 1], 1 / (1 - pi_hat[A == 0])),
  A      = factor(c(rep(1, sum(A == 1)), rep(0, sum(A == 0))),
                  labels = c("No abciximab", "Abciximab"))
) %>%
  ggplot(aes(x = A, y = weight, fill = A)) +
  geom_boxplot(alpha = 0.6, outlier.shape = 21) +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  theme_minimal() +
  labs(x = NULL, y = "Inverse weight",
       fill  = "Treatment",
       title = "Distribution of inverse weights by treatment group (no cross-fitting)")

# ── Horvitz-Thompson estimator ────────────────────────────────────────────────
ipw_weights <- A / pi_hat - (1 - A) / (1 - pi_hat)
ate_ipw     <- mean(ipw_weights * Y)
cat("IPW ATE - Horvitz-Thompson (no cross-fitting):", round(ate_ipw, 4), "\n")

# ── Weighted linear regression ────────────────────────────────────────────────
weights_sl  <- A / pi_hat + (1 - A) / (1 - pi_hat)
fit_ipw_sl  <- lm(Y ~ A, weights = weights_sl)

ate_ipw_sl    <- coef(fit_ipw_sl)["A"]
ate_ipw_sl_se <- sqrt(vcovHC(fit_ipw_sl, type = "HC")["A", "A"])
ci_ipw_sl     <- ate_ipw_sl + c(-1, 1) * 1.96 * ate_ipw_sl_se

cat("IPW ATE - weighted regression (no cross-fitting):", round(ate_ipw_sl, 4), "\n")
cat("SE:", round(ate_ipw_sl_se, 4), "\n")
cat("95% CI: [", round(ci_ipw_sl[1], 4), ",", round(ci_ipw_sl[2], 4), "]\n")

# ── With cross-fitting ────────────────────────────────────────────────────────
pi_hat_cf <- numeric(N)

for (k in 1:n_folds) {

  test  <- folds[[k]]
  train <- setdiff(seq_len(N), test)

  sl_pi_cf <- SuperLearner(
    Y          = A[train],
    X          = X[train, ],
    SL.library = SL.library,
    family     = binomial(),
    method     = "method.NNLS",
    cvControl  = list(V = 5)
  )

  pi_hat_cf[test] <- predict(sl_pi_cf, newdata = X[test, ])$pred
}

# ── Positivity check ──────────────────────────────────────────────────────────
summary(pi_hat_cf)

data.frame(pi_hat_cf = pi_hat_cf, A = factor(A, labels = c("No abciximab", "Abciximab"))) %>%
  ggplot(aes(x = pi_hat_cf, fill = A, color = A)) +
  geom_density(alpha = 0.3, linewidth = 0.8) +
  scale_fill_manual(values  = c("steelblue", "tomato")) +
  scale_color_manual(values = c("steelblue", "tomato")) +
  theme_minimal() +
  labs(x     = expression(hat(pi)(X)),
       y     = "Density",
       fill  = "Treatment",
       color = "Treatment",
       title = "Propensity scores by treatment group (cross-fitting)")

data.frame(
  weight = c(1 / pi_hat_cf[A == 1], 1 / (1 - pi_hat_cf[A == 0])),
  A      = factor(c(rep(1, sum(A == 1)), rep(0, sum(A == 0))),
                  labels = c("No abciximab", "Abciximab"))
) %>%
  ggplot(aes(x = A, y = weight, fill = A)) +
  geom_boxplot(alpha = 0.6, outlier.shape = 21) +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  theme_minimal() +
  labs(x = NULL, y = "Inverse weight",
       fill  = "Treatment",
       title = "Distribution of inverse weights by treatment group (cross-fitting)")

# ── Horvitz-Thompson estimator ────────────────────────────────────────────────
ipw_weights_cf <- A / pi_hat_cf - (1 - A) / (1 - pi_hat_cf)
ate_ipw_cf     <- mean(ipw_weights_cf * Y)
cat("IPW ATE - Horvitz-Thompson (cross-fitting):", round(ate_ipw_cf, 4), "\n")

# ── Weighted linear regression ────────────────────────────────────────────────
weights_sl_cf  <- A / pi_hat_cf + (1 - A) / (1 - pi_hat_cf)
fit_ipw_sl_cf  <- lm(Y ~ A, weights = weights_sl_cf)

ate_ipw_cf_sl    <- coef(fit_ipw_sl_cf)["A"]
ate_ipw_cf_sl_se <- sqrt(vcovHC(fit_ipw_sl_cf, type = "HC")["A", "A"])
ci_ipw_cf_sl     <- ate_ipw_cf_sl + c(-1, 1) * 1.96 * ate_ipw_cf_sl_se

cat("IPW ATE - weighted regression (cross-fitting):", round(ate_ipw_cf_sl, 4), "\n")
cat("SE:", round(ate_ipw_cf_sl_se, 4), "\n")
cat("95% CI: [", round(ci_ipw_cf_sl[1], 4), ",", round(ci_ipw_cf_sl[2], 4), "]\n")
```

## 6. AIPW

The Augmented IPW (AIPW) estimator combines the outcome model and the
propensity score into a single *doubly robust* estimator: it remains
consistent if either (but not necessarily both) nuisance models are
correctly specified.

$$
\widehat{\text{ATE}}_{\text{AIPW}} = \frac{1}{n} \sum_{i=1}^{n}
\left[
  \hat{Q}(1, X_i) - \hat{Q}(0, X_i)
  + \frac{A_i (Y_i - \hat{Q}(1, X_i))}{\hat{\pi}(X_i)}
  - \frac{(1 - A_i)(Y_i - \hat{Q}(0, X_i))}{1 - \hat{\pi}(X_i)}
\right]
$$

## 6.1 npcausal AIPW

An easy way to compute the AIPW estimator with cross-fitting is to use
the [`npcausal`](https://rdrr.io/github/ehkennedy/npcausal/) package. To
install it from GitHub:

```{r installation}
#install.packages("devtools")
#library(devtools)
#install_github("ehkennedy/npcausal")
library(npcausal)
```

The `ate()` function handles cross-fitting and inference internally,
using the same number of folds and learner library as our manual
implementation.

```{r npcausal-aipw}

ate_npcausal <- ate(
  y       = Y,
  a       = A,
  x       = X,
  nsplits = 5,
  sl.lib  = SL.library
)

print(ate_npcausal$res)
```

### 6.2 manual AIPW

```{r manual aipw}
# ── Without cross-fitting ─────────────────────────────────────────────────────
aug_1    <-  A      * (Y - Q1) / pi_hat
aug_0    <- (1 - A) * (Y - Q0) / (1 - pi_hat)
psi_i    <- (Q1 - Q0) + aug_1 - aug_0
ate_aipw <- mean(psi_i)
se_aipw  <- sd(psi_i) / sqrt(N)
ci_aipw  <- ate_aipw + c(-1, 1) * 1.96 * se_aipw

cat("AIPW ATE (no cross-fitting):", round(ate_aipw, 4), "\n")
cat("95% CI: [", round(ci_aipw[1], 4), ",", round(ci_aipw[2], 4), "]\n")

# ── With cross-fitting ────────────────────────────────────────────────────────
aug_1_cf    <-  A      * (Y - Q1_cf) / pi_hat_cf
aug_0_cf    <- (1 - A) * (Y - Q0_cf) / (1 - pi_hat_cf)
psi_i_cf    <- (Q1_cf - Q0_cf) + aug_1_cf - aug_0_cf
ate_aipw_cf <- mean(psi_i_cf)
se_aipw_cf  <- sd(psi_i_cf) / sqrt(N)
ci_aipw_cf  <- ate_aipw_cf + c(-1, 1) * 1.96 * se_aipw_cf

cat("AIPW ATE (cross-fitting):", round(ate_aipw_cf, 4), "\n")
cat("95% CI: [", round(ci_aipw_cf[1], 4), ",", round(ci_aipw_cf[2], 4), "]\n")
```

## 7. TMLE

Targeted Maximum Likelihood Estimation (TMLE) is a doubly robust,
asymptotically efficient substitution estimator. Rather than directly
correcting the plug-in bias as AIPW does, TMLE updates the initial
outcome model \$\\hat{Q}\$ through a targeted fluctuation step driven by
the propensity score, ensuring the final estimate solves the efficient
score equation exactly. We illustrate two implementations: the [\`tmle\`
package](https://cran.r-project.org/web/packages/tmle/refman/tmle.html)
and a manual derivation that makes the fluctuation step explicit.

### 7.1 tmle Package

```{r tmle}
# ── Without cross-fitting ─────────────────────────────────────────────────────
tmle_fit <- tmle(
  Y            = Y,
  A            = A,
  W            = as.data.frame(X),
  Q.SL.library = SL.library,
  g.SL.library = SL.library,
  family       = "binomial",
  V.Q          = 1,
  V.g          = 1
)

ate_tmle <- tmle_fit$estimates$ATE$psi
ci_tmle  <- tmle_fit$estimates$ATE$CI

cat("TMLE ATE (no cross-fitting):", round(ate_tmle, 4), "\n")
cat("95% CI: [", round(ci_tmle[1], 4), ",", round(ci_tmle[2], 4), "]\n")

# ── With cross-fitting ────────────────────────────────────────────────────────
# Supply cross-fitted nuisance estimates directly via Q and g
tmle_fit_cf <- tmle(
  Y            = Y,
  A            = A,
  W            = as.data.frame(X),
  Q.SL.library = SL.library,
  g.SL.library = SL.library,
  family       = "binomial",
  Q            = cbind(Q0_cf, Q1_cf),
  g1W          = pi_hat_cf
)

ate_tmle_cf <- tmle_fit_cf$estimates$ATE$psi
ci_tmle_cf  <- tmle_fit_cf$estimates$ATE$CI

cat("TMLE ATE (cross-fitting):", round(ate_tmle_cf, 4), "\n")
cat("95% CI: [", round(ci_tmle_cf[1], 4), ",", round(ci_tmle_cf[2], 4), "]\n")
```

### 7.2 manual TMLE

```{r manual tmle}
# ── Clever Covariates ─────────────────────────────────────────────────────────
H1 <- A / pi_hat_cf
H0 <- (1 - A) / (1 - pi_hat_cf)

# ── Estimate fluctuation parameters ──────────────────────────────────────────
delta1 <- coef(glm(Y ~ -1 + offset(qlogis(Q1_cf)) + H1, family = binomial()))
delta0 <- coef(glm(Y ~ -1 + offset(qlogis(Q0_cf)) + H0, family = binomial()))

# ── Update initial predictions ────────────────────────────────────────────────
Q1_cf_updated <- plogis(qlogis(Q1_cf) + delta1 / pi_hat_cf)
Q0_cf_updated <- plogis(qlogis(Q0_cf) + delta0 / (1 - pi_hat_cf))

# ── ATE estimate ──────────────────────────────────────────────────────────────
ate_tmle_cf_manual <- mean(Q1_cf_updated - Q0_cf_updated)

# ── Efficient Influence Function and standard error ───────────────────────────
eif <- (Q1_cf_updated - Q0_cf_updated - ate_tmle_cf_manual) +
        A / pi_hat_cf * (Y - Q1_cf_updated) -
        (1 - A) / (1 - pi_hat_cf) * (Y - Q0_cf_updated)

se_tmle_cf_manual  <- sd(eif) / sqrt(N)
ci_tmle_cf_manual  <- ate_tmle_cf_manual + c(-1, 1) * 1.96 * se_tmle_cf_manual

cat("TMLE ATE (cross-fitting, manual):", round(ate_tmle_cf_manual, 4), "\n")
cat("SE:", round(se_tmle_cf_manual, 4), "\n")
cat("95% CI: [", round(ci_tmle_cf_manual[1], 4), ",", round(ci_tmle_cf_manual[2], 4), "]\n")
```

## 8. Comparison of Estimators

```{r comparison}
results <- tibble(
  Estimator    = c("G-Computation", "G-Computation",
                   "IPW",           "IPW",
                   "AIPW",          "AIPW",
                   "TMLE",          "TMLE"),
  Cross_fitting = rep(c("No", "Yes"), 4),
  ATE          = c(ate_gcomp,    ate_gcomp_cf,
                   ate_ipw,      ate_ipw_cf,
                   ate_aipw,     ate_aipw_cf,
                   ate_tmle,     ate_tmle_cf),
  CI_lower     = c(NA, NA, NA, NA,
                   ci_aipw[1],    ci_aipw_cf[1],
                   ci_tmle[1],    ci_tmle_cf[1]),
  CI_upper     = c(NA, NA, NA, NA,
                   ci_aipw[2],    ci_aipw_cf[2],
                   ci_tmle[2],    ci_tmle_cf[2])
)

print(results)

# ── Forest-style plot ─────────────────────────────────────────────────────────
results %>%
  mutate(Estimator = factor(Estimator, levels = rev(c("G-Computation", "IPW", "AIPW", "TMLE")))) %>%
  ggplot(aes(x = ATE, y = Estimator, color = Cross_fitting, shape = Cross_fitting)) +
  geom_point(size = 3, position = position_dodge(width = 0.4)) +
  geom_errorbarh(aes(xmin = CI_lower, xmax = CI_upper),
                 height = 0.2, linewidth = 0.8,
                 position = position_dodge(width = 0.4),
                 na.rm = TRUE) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = c("steelblue", "tomato")) +
  theme_minimal() +
  labs(x = "Estimated ATE", y = NULL, color = "Cross-fitting", shape = "Cross-fitting",
       title = "ATE estimates across estimators and cross-fitting strategies")
```

## (Bonus) 9. AIPW and TMLE for ATT

In this question we will repeat the same analysis for the average
treatment effect on th treated, which is defined as:

$$
ATT = E[Y^1 - Y^0 \mid A=1],
$$

which can be nonparametrically identified from observed data as:

$$
\text{ATT} = E\Big[E[Y \mid A=1, X] - E[Y \mid A=0, X]\mid A=1\Big].
$$

### 9.1 AIPW for ATT

```{r npcausal ATT}

att_npcausal <- att(
  y       = Y,
  a       = A,
  x       = X,
  nsplits = 5,
  sl.lib  = SL.library
)

print(att_npcausal$res)
```

```{r estimating equations ATT}

att_cf_manual <- mean((A/mean(A))*(Y-Q0_cf)-((1-A)*pi_hat_cf/(1-pi_hat_cf))*(Y-Q0_cf)/mean(A))

att_ic <- (A/mean(A))*(Y-Q0_cf-att_cf_manual)-((1-A)*pi_hat_cf/(1-pi_hat_cf))*(Y-Q0_cf)/mean(A)

c(att_cf_manual,sd(att_ic)/sqrt(length(A)))
```

### 9.2 TMLE for ATT

```{r tmle ATT}

att_tmle_cf <- tmle_fit_cf$estimates$ATT$psi
ci_att_tmle_cf  <- tmle_fit_cf$estimates$ATT$CI

cat("TMLE ATT (cross-fitting):", round(att_tmle_cf, 4), "\n")
cat("95% CI: [", round(ci_tmle_cf[1], 4), ",", round(ci_att_tmle_cf[2], 4), "]\n")
```

```{r manual TMLE ATT}

fluctmod_att <- lm(Y[A==0]~-1+offset(Q0_cf[A==0])+I(pi_hat_cf[A==0]/(1-pi_hat_cf[A==0])))

Q0_cf_updated_att <- Q0_cf + fluctmod_att$coef*pi_hat_cf/(1-pi_hat_cf)

att_tmle_cf_manual <- mean((A/mean(A))*(Y-Q0_cf_updated_att))

att_ic_tmle <- (A/mean(A))*(Y-Q0_cf_updated_att-att_tmle_cf_manual)-((1-A)*pi_hat_cf/(1-pi_hat_cf))*(Y-Q0_cf_updated_att)/mean(A) 

c(att_tmle_cf_manual,sd(att_ic_tmle)/sqrt(length(A)))
```
