---
title: "Lab 9: Instrumental Variable Analysis"
subtitle: "Causal Machine Learning"
author: "Stijn Vansteelandt and Federico Bertoia, Ghent University"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## 0. Introduction

The goal of this lab is to perform **Instrumental Variable (IV)
analysis** using data from a study of 139 premenopausal women. We are
interested in estimating the causal effect of serum IGFBP-3 levels on
total dense breast tissue area, while accounting for unmeasured
confounding.

Standard regression methods require that all common causes of the
exposure and outcome are measured and adjusted for — an assumption that
is often untestable and likely violated in observational data. When
unmeasured confounders $U$ are present, IV methods can still identify
causal effects, provided a valid instrument $R$ is available.

An **instrumental variable** $R$ must satisfy three assumptions:

1.  **Relevance**: $R$ is associated with the exposure $A$,
2.  **Exclusion restriction**: $R$ affects the outcome $Y$ only through
    $A$,
3.  **Randomization**: $R$ is (conditionally) independent of $$Y^a$$.

The **estimand** under a linear structural mean model
$E(Y \mid A, L, R, U) = \psi A + g(L, U)$ is:

$$
\psi = \frac{E\{\text{Cov}(R, Y \mid L)\}}{E\{\text{Cov}(R, A \mid L)\}}
$$

This can be estimated via the **Efficient Influence Function (EIF)**
estimator:

$$
\hat{\psi} =
\frac{\sum_{i=1}^n \bigl\{R_i - \hat{E}(R_i \mid L_i)\bigr\}
\bigl\{Y_i - \hat{E}(Y_i \mid L_i)\bigr\}}
{\sum_{i=1}^n \bigl\{R_i - \hat{E}(R_i \mid L_i)\bigr\}
\bigl\{A_i - \hat{E}(A_i \mid L_i)\bigr\}}
$$

with standard error estimated as $n^{-1/2}$ times the sample standard
deviation of the influence function values divided by the denominator.
Cross-fitting with data-adaptive nuisance estimators (Super Learner) is
recommended.

------------------------------------------------------------------------

### Data

The dataset `igf.dta` contains observations on 139 premenopausal women.

| Variable  | Role           | Description                               |
|-----------|----------------|-------------------------------------------|
| `lucent`  | Outcome $Y$    | Total area of dense tissue in the breast  |
| `igfbp3`  | Exposure $A$   | Serum IGFBP-3 levels                      |
| `igfbp3g` | Instrument $R$ | Putative functional genotype (AA, AC, CC) |
| `age`     | Confounder $L$ | Patient's age                             |
| `afbcat`  | Confounder $L$ | Age at first birth (categorized)          |
| `brfcat`  | Confounder $L$ | Breast feeding duration                   |
| `parcat`  | Confounder $L$ | Number of children                        |
| `ev_pil`  | Confounder $L$ | Ever taken oral contraceptives (binary)   |
| `bmi`     | —              | Body mass index                           |
| `ev_smok` | —              | Ever smoked (binary)                      |

Both the exposure and outcome will be log-transformed throughout to
facilitate interpretation and improve distributional symmetry.

```{r setup-data}
# ── Packages ──────────────────────────────────────────────────────────────────
library(foreign)      # read.dta
library(dplyr)
library(ggplot2)
library(patchwork)
library(tibble)
library(SuperLearner)
library(lmtest)
library(sandwich)

# ── Reproducibility ───────────────────────────────────────────────────────────
set.seed(123)

# ── Load data ─────────────────────────────────────────────────────────────────
data <- read.dta("igf.dta")
N    <- nrow(data)

# ── Derived variables ─────────────────────────────────────────────────────────
data <- data %>%
  mutate(
    log_lucent = log(lucent),
    log_igfbp3 = log(igfbp3)
  )

glimpse(data)
```

------------------------------------------------------------------------

## 1. Question 1 — Association of Lucent Area with Genotype

We begin by testing whether the proposed instrument `igfbp3g` is
associated with the outcome `log(lucent)`. Under the IV assumptions, any
such association can only arise through the exposure `igfbp3`. A
significant association is therefore consistent with (but does not
prove) the existence of a causal effect of IGFBP-3 on breast density.

```{r q1-outcome-iv-association}
# ── Numeric coding of genotype ────────────────────────────────────────────────
# AA = 1, AC = 2, CC = 3 (allele dose ordering)
data <- data %>%
  mutate(igfbp3g_num = as.numeric(igfbp3g))

# ── Linear regression of log(lucent) on genotype (linear trend) ───────────────
lm_q1 <- lm(log_lucent ~ igfbp3g_num, data = data)
summary(lm_q1)

# ── As a factor (no ordering assumed) ────────────────────────────────────────
lm_q1_cat <- lm(log_lucent ~ factor(igfbp3g), data = data)
summary(lm_q1_cat)

# ── Pearson correlation ───────────────────────────────────────────────────────
cor.test(data$log_lucent, data$igfbp3g_num)
```

> **Interpretation:** The genotype is significantly associated with
> log(lucent area) ($p = 0.018$), with each additional C allele
> corresponding to a decrease in log-lucent area of approximately 0.154.
> Under the IV assumptions, this association propagates entirely through
> IGFBP-3 levels — consistent with the instrument having a causal
> downstream effect.

------------------------------------------------------------------------

## 2. Question 2 — Distribution of IGFBP-3 and Relevance of the Instrument

We examine the distribution of the exposure and assess whether the
instrument satisfies the **relevance** assumption — i.e., whether
`igfbp3g` is associated with serum IGFBP-3 after (implicitly) adjusting
for measured confounders. This is the only IV assumption directly
testable from data.

### 2.1 Histogram of IGFBP-3 (raw and log-transformed)

```{r q2-histograms}
p1 <- ggplot(data, aes(x = igfbp3)) +
  geom_histogram(bins = 20, fill = "steelblue", alpha = 0.7, color = "white") +
  theme_minimal() +
  labs(x = "igfbp3", y = "Frequency", title = "Histogram of igfbp3")

p2 <- ggplot(data, aes(x = log_igfbp3)) +
  geom_histogram(bins = 20, fill = "steelblue", alpha = 0.7, color = "white") +
  theme_minimal() +
  labs(x = "log_igfbp3", y = "Frequency", title = "Histogram of log_igfbp3")

p1 + p2
```

> The raw distribution of IGFBP-3 is right-skewed; the log
> transformation yields an approximately symmetric distribution, making
> it more suitable for linear models.

### 2.2 Distribution of log(IGFBP-3) by genotype

```{r q2-by-genotype}
# ── Summary statistics by genotype ────────────────────────────────────────────
data %>%
  group_by(igfbp3g) %>%
  summarise(
    n    = n(),
    mean = round(mean(log_igfbp3), 3),
    sd   = round(sd(log_igfbp3), 3),
    min  = round(min(log_igfbp3), 3),
    max  = round(max(log_igfbp3), 3)
  )

# ── Overlapping histograms ────────────────────────────────────────────────────
ggplot(data, aes(x = log_igfbp3, fill = igfbp3g, colour = igfbp3g)) +
  geom_histogram(alpha = 0.4, position = "identity", bins = 20) +
  scale_fill_manual(values  = c("tomato", "steelblue", "forestgreen")) +
  scale_colour_manual(values = c("tomato", "steelblue", "forestgreen")) +
  theme_minimal() +
  labs(x = "log_igfbp3", y = "Count",
       title = "Distribution of log(IGFBP-3) by genotype",
       fill = "igfbp3g", colour = "igfbp3g")
```

### 2.3 Testing relevance

```{r q2-relevance}
# ── Linear regression of log(igfbp3) on genotype ─────────────────────────────
lm_q2 <- lm(log_igfbp3 ~ igfbp3g_num, data = data)
summary(lm_q2)

# ── As factor, with F-test ────────────────────────────────────────────────────
lm_q2_cat <- lm(log_igfbp3 ~ factor(igfbp3g), data = data)
anova(lm_q2_cat)
```

> **Interpretation:** The genotype is significantly associated with
> log(IGFBP-3) ($p = 0.009$), supporting the relevance assumption. Each
> additional C allele is associated with a decrease of approximately
> 0.063 in log(IGFBP-3). This is the only IV assumption directly
> testable from data; the exclusion restriction and randomization
> assumption require subject-matter justification.

------------------------------------------------------------------------

## 3. Question 3 — DAG and IV Assumptions

```{r q3-dag}
library(DiagrammeR)

grViz("
digraph IV_DAG {
  graph [layout = neato, bgcolor = transparent, splines = curved]
  node [shape = plaintext, fontname = 'Helvetica', fontsize = 16]
  R [pos = '0,0!',  label = 'R\n(igfbp3g)']
  A [pos = '3,0!',  label = 'A\n(log igfbp3)']
  Y [pos = '6,0!',  label = 'Y\n(log lucent)']
  U [pos = '4.5,2!', label = 'U']
  L [pos = '3,-2!', label = 'L\n(age, afbcat,\nbrfcat, parcat, ev_pil)']

  edge [arrowhead = vee, arrowsize = 0.8, penwidth = 1.6, color = '#444444']
  R -> A
  A -> Y
  U -> A
  U -> Y
  L -> A
  L -> Y

  edge [arrowhead = vee, arrowsize = 0.7, penwidth = 1.2,
        style = dashed, color = '#888888']
  R -> Y [style = dashed, constraint = false]
  R -> L [style = dashed]
}
")
```

The three IV assumptions in this setting are:

1.  **Relevance** ($R \not\!\perp\!\!\!\perp A \mid L$): The IGFBP3g
    genotype must be associated with serum IGFBP-3 levels after
    adjusting for observed covariates $L$. This was verified empirically
    in Question 2.

2.  **Exclusion restriction** ($R \perp\!\!\!\perp Y \mid A, L, U$): The
    genotype should not affect dense breast tissue area through any
    pathway other than serum IGFBP-3. This assumption is *potentially
    violated* if the genotype affects fertility-related traits (such as
    parity or age at first birth), which in turn influence breast tissue
    density.

3.  **Randomization** ($R \perp\!\!\!\perp U \mid L$): The genotype is
    randomized by Mendelian inheritance — transmission of alleles from
    parent to offspring is random, so it should not share common causes
    with the unmeasured confounders $U$ after conditioning on $L$. This
    assumption is not testable from data and relies on biological
    reasoning.

------------------------------------------------------------------------

## 4. Question 4 — IV Estimate via Efficient Influence Function (no covariates)

We estimate $\psi$ using the EIF estimator without adjusting for any
covariates $L$. In this case $\hat{E}(R \mid L) = \bar{R}$,
$\hat{E}(A \mid L) = \bar{A}$, and $\hat{E}(Y \mid L) = \bar{Y}$, so the
estimator reduces to:

$$
\hat{\psi} =
\frac{\sum_i (R_i - \bar{R})(Y_i - \bar{Y})}
     {\sum_i (R_i - \bar{R})(A_i - \bar{A})}
$$

```{r q4-iv-no-covariates}
# ── Variables ─────────────────────────────────────────────────────────────────
Y <- data$log_lucent
A <- data$log_igfbp3
R <- data$igfbp3g_num

# ── EIF estimator (no covariates) ─────────────────────────────────────────────
psi <- mean((R - mean(R)) * (Y - mean(Y))) /
       mean((R - mean(R)) * (A - mean(A)))

# ── Standard error via influence function ─────────────────────────────────────
EIC_psi <- (R - mean(R)) * (Y - mean(Y) - psi * (A - mean(A))) /
           mean((R - mean(R)) * (A - mean(A)))
se_psi <- sd(EIC_psi) / sqrt(N)

# ── 95% confidence interval ───────────────────────────────────────────────────
psi_lb <- psi - 1.96 * se_psi
psi_ub <- psi + 1.96 * se_psi

# ── Results table (on geometric mean ratio scale for a 10% increase in A) ─────
tibble(
  Method    = "Eff. Influence Function",
  Estimate  = format(round(exp(psi     * log(1.1)), 3), nsmall = 3),
  CI_Lower  = format(round(exp(psi_lb  * log(1.1)), 3), nsmall = 3),
  CI_Upper  = format(round(exp(psi_ub  * log(1.1)), 3), nsmall = 3)
)
```

> **Interpretation:** A 10% increase in serum IGFBP-3 increases the
> geometric mean of dense breast tissue area by approximately 26.1%
> ($= \exp(\hat\psi \cdot \log 1.1)$) [95% CI: 0.1% to 58.9%]. The
> confidence interval is wide, reflecting the limited sample size and
> weak instrument strength.

------------------------------------------------------------------------

## 5. Question 5 — DML IV Estimate with Covariates (SuperLearner)

A concern with the unadjusted IV estimate is that `parcat` and `afbcat`
(parity and age at first birth) may lie on the causal pathway from
genotype to outcome — violating the exclusion restriction. We address
this by conditioning on $L = (\texttt{afbcat}, \texttt{parcat})$,
thereby blocking that pathway.

We use **Debiased Machine Learning** with SuperLearner to estimate the
three nuisance functions $\hat{E}(Y \mid L)$, $\hat{E}(A \mid L)$ and
$\hat{E}(R \mid L)$. Given the small sample ($n = 139$), cross-fitting
is omitted and a restricted library is used.

```{r q5-dml-iv}
# ── Confounder set blocking the problematic pathway ───────────────────────────
L <- data %>% dplyr::select(afbcat, parcat)

SL.library <- c("SL.glm", "SL.ranger")

# ── Nuisance models ───────────────────────────────────────────────────────────
sl_Y <- SuperLearner(Y = Y, X = L, SL.library = SL.library)
sl_A <- SuperLearner(Y = A, X = L, SL.library = SL.library)
sl_R <- SuperLearner(Y = R, X = L, SL.library = SL.library)

# ── Nuisance predictions ──────────────────────────────────────────────────────
m_Y <- predict(sl_Y, newdata = L, onlySL = TRUE)$pred
m_A <- predict(sl_A, newdata = L, onlySL = TRUE)$pred
m_R <- predict(sl_R, newdata = L, onlySL = TRUE)$pred

# ── EIF estimator with covariates ─────────────────────────────────────────────
psi_L <- mean((R - m_R) * (Y - m_Y)) /
         mean((R - m_R) * (A - m_A))

# ── Standard error ────────────────────────────────────────────────────────────
EIC_psi_L <- (R - m_R) * (Y - m_Y - psi_L * (A - m_A)) /
             mean((R - m_R) * (A - m_A))
se_psi_L  <- sd(EIC_psi_L) / sqrt(N)

# ── 95% confidence interval ───────────────────────────────────────────────────
psi_L_lb <- psi_L - 1.96 * se_psi_L
psi_L_ub <- psi_L + 1.96 * se_psi_L

# ── Results table ─────────────────────────────────────────────────────────────
tibble(
  Method   = "Eff. Influence Function",
  Estimate = format(round(exp(psi_L    * log(1.1)), 3), nsmall = 3),
  CI_Lower = format(round(exp(psi_L_lb * log(1.1)), 3), nsmall = 3),
  CI_Upper = format(round(exp(psi_L_ub * log(1.1)), 3), nsmall = 3)
)
```

> **Interpretation:** After conditioning on `afbcat` and `parcat` to
> block the potentially problematic pathway from genotype to outcome, a
> 10% increase in serum IGFBP-3 increases the geometric mean of dense
> breast tissue area by approximately 24.9% [95% CI: 3.4% to 50.9%]. The
> estimate is slightly attenuated relative to Question 4 and the
> interval no longer includes zero.

------------------------------------------------------------------------

## 6. Question 6 — Assumption-Lean Regression (Comparison)

For comparison, we estimate the **association** (not necessarily causal
effect) between log(IGFBP-3) and log(lucent area) using
**assumption-lean linear regression** — adjusting for the full
confounder set $L = (\texttt{age}, \texttt{afbcat}, \texttt{brfcat},
\texttt{parcat}, \texttt{ev\_pil})$. We use two equivalent approaches:

-   The **EIF estimator** based on the partially linear model
    $E(Y \mid A, L) = \beta A + g(L)$, which targets the coefficient
    $\beta = E[(A - E(A \mid L))^2]^{-1} \cdot E[(A - E(A \mid L))(Y - E(Y \mid L))]$.
-   The **OLS residual-on-residual regression** with
    heteroskedasticity-robust (sandwich) standard errors.

```{r q6-assumption-lean}
# ── Full confounder set ───────────────────────────────────────────────────────
L2 <- data %>% dplyr::select(age, afbcat, brfcat, parcat, ev_pil)

# ── Nuisance models ───────────────────────────────────────────────────────────
sl_Y2 <- SuperLearner(Y = Y, X = L2, SL.library = SL.library)
sl_A2 <- SuperLearner(Y = A, X = L2, SL.library = SL.library)

m_Y2 <- predict(sl_Y2, newdata = L2, onlySL = TRUE)$pred
m_A2 <- predict(sl_A2, newdata = L2, onlySL = TRUE)$pred

# ── EIF estimator ─────────────────────────────────────────────────────────────
beta_eif <- mean((A - m_A2) * (Y - m_Y2)) /
            mean((A - m_A2)^2)

EIC_beta  <- (A - m_A2) * (Y - m_Y2 - beta_eif * (A - m_A2)) /
             mean((A - m_A2)^2)
se_eif    <- sd(EIC_beta) / sqrt(N)

beta_eif_lb <- beta_eif - 1.96 * se_eif
beta_eif_ub <- beta_eif + 1.96 * se_eif

# ── OLS with sandwich SE ──────────────────────────────────────────────────────
lm_resid  <- lm(I(Y - m_Y2) ~ -1 + I(A - m_A2))
lm_result <- coeftest(lm_resid, vcov = sandwich)

beta_lm    <- lm_result[1, 1]
se_lm      <- lm_result[1, 2]
beta_lm_lb <- beta_lm - 1.96 * se_lm
beta_lm_ub <- beta_lm + 1.96 * se_lm

# ── Combined results table ────────────────────────────────────────────────────
tibble(
  Method    = c("Linear Model", "Eff. Influence Function"),
  Estimate  = format(round(exp(c(beta_lm,    beta_eif)    * log(1.1)), 3), nsmall = 3),
  CI_Lower  = format(round(exp(c(beta_lm_lb, beta_eif_lb) * log(1.1)), 3), nsmall = 3),
  CI_Upper  = format(round(exp(c(beta_lm_ub, beta_eif_ub) * log(1.1)), 3), nsmall = 3)
)
```

> **Interpretation:** The assumption-lean regression estimates a much
> smaller association: a 10% increase in IGFBP-3 is associated with a
> 4.5% increase in the geometric mean of dense breast tissue area [95%
> CI: −0.2% to 9.3%], which is not statistically significant. This
> contrasts sharply with the IV estimates (\~25%), suggesting that
> unmeasured confounding substantially attenuates the observational
> association.

------------------------------------------------------------------------

## 7. Summary of Results

```{r summary-table}
results_all <- tibble(
  Method = c(
    "IV — EIF (no covariates)",
    "IV — DML with covariates (SuperLearner)",
    "Assumption-lean regression — LM",
    "Assumption-lean regression — EIF"
  ),
  Estimate = format(round(exp(c(psi, psi_L, beta_lm, beta_eif) * log(1.1)), 3), nsmall = 3),
  CI_Lower = format(round(exp(c(psi_lb,    psi_L_lb, beta_lm_lb, beta_eif_lb) * log(1.1)), 3), nsmall = 3),
  CI_Upper = format(round(exp(c(psi_ub,    psi_L_ub, beta_lm_ub, beta_eif_ub) * log(1.1)), 3), nsmall = 3)
)

print(results_all)
```

```{r forest-plot}
tibble(
  Method   = c(
    "IV — EIF (no covariates)",
    "IV — DML with covariates",
    "Assumption-lean (LM)",
    "Assumption-lean (EIF)"
  ),
  Estimate = exp(c(psi, psi_L, beta_lm, beta_eif) * log(1.1)),
  Lower    = exp(c(psi_lb, psi_L_lb, beta_lm_lb, beta_eif_lb) * log(1.1)),
  Upper    = exp(c(psi_ub, psi_L_ub, beta_lm_ub, beta_eif_ub) * log(1.1))
) %>%
  mutate(Method = factor(Method, levels = rev(Method))) %>%
  ggplot(aes(x = Estimate, y = Method)) +
  geom_point(size = 3, color = "tomato") +
  geom_errorbarh(aes(xmin = Lower, xmax = Upper),
                 height = 0.25, linewidth = 0.8, color = "tomato") +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey50") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1,
                                                      scale = 100)) +
  theme_minimal() +
  labs(x = "Geometric mean ratio (10% increase in IGFBP-3)",
       y   = NULL,
       title = "IV and regression estimates — IGFBP-3 effect on breast density")
```

> **Overall conclusion:** The IV estimates suggest a substantial
> positive causal effect of serum IGFBP-3 on dense breast tissue area,
> with a 10% increase in IGFBP-3 associated with roughly a 25% increase
> in the geometric mean lucent area. The assumption-lean regression
> estimate is substantially smaller and non-significant, pointing to
> negative unmeasured confounding that masks the causal effect in a
> standard observational analysis. The IV confidence intervals are wide
> due to the modest instrument strength and small sample, so results
> should be interpreted with caution.
