---
title: "Lab 8: Assumption-Lean Regression"
subtitle: "Causal Machine Learning"
author: "Stijn Vansteelandt and Federico Bertoia, Ghent University"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## 0. Introduction

The goal of this lab is to apply **assumption-lean regression** to
estimate the effect of a health program on a binary outcome, using the
King County 2001 birth data. The key idea is to use a semiparametric
model as an **approximation** rather than an assumption, and to estimate
the resulting model-free estimand via its efficient influence function
combined with machine learning.

### Motivation

Consider the partially linear model

$$
E(Y^a \mid L) = \alpha(L) + \beta(L) \cdot a.
$$

Instead of assuming $\beta(L) = \beta$ (constant), we summarize the
distribution of $\beta(L)$ by its **overlap-weighted average**:

$$
\beta = \frac{E\{\text{Cov}(A, Y \mid L)\}}{E\{\text{Var}(A \mid L)\}} = \frac{E\{w(L)\,\beta(L)\}}{E\{w(L)\}},
$$

where for a binary exposure $A$, the overlap weights are
$w(L) = P(A=1 \mid L)\,P(A=0 \mid L)$ and
$\beta(L) = E(Y \mid A=1, L) - E(Y \mid A=0, L)$.

The **efficient influence curve** of $\beta$ is:

$$
\mathrm{IF}(\beta) = \frac{\{A - E(A \mid L)\}\bigl[Y - E(Y \mid L) - \beta\{A - E(A \mid L)\}\bigr]}{E\bigl[\{A - E(A \mid L)\}^2\bigr]},
$$

which leads to the debiased estimator:

$$
\hat{\beta} = \frac{\sum_{i=1}^n \bigl\{A_i - \hat{E}(A_i \mid L_i)\bigr\}\bigl\{Y_i - \hat{E}(Y_i \mid L_i)\bigr\}}{\sum_{i=1}^n \bigl\{A_i - \hat{E}(A_i \mid L_i)\bigr\}^2}.
$$

In practice this corresponds to linearly regressing $(Y - \hat{q})$ onto
$(A - \hat{p})$ with no intercept, using a sandwich (robust) standard
error. Cross-fitting is recommended to avoid overfitting bias.

------------------------------------------------------------------------

### Data

The **King County 2001** dataset contains birth certificate records from
2,500 singleton births in King County, Washington (2001). The program of
interest is **First Steps**, a 1989 Washington State initiative aimed at
improving maternal and infant health for low-income pregnant women.

| Variable | Role | Description |
|----------------|----------------|-----------------------------------------|
| `firstep` | Treatment | Participation in First Steps program (binary) |
| `low_bwt` | Outcome | Low birth weight: `bwt < 2500` grams (binary) |
| `age` | Exposure / Confounder | Maternal age in years |
| `gender` | Confounder | Child's gender |
| `race` | Confounder | Maternal race |
| `parity` | Confounder | Number of previous live-born infants |
| `married` | Confounder | Marital status |
| `smoker` | Post-treatment | Smoking during pregnancy (excluded) |
| `drinker` | Post-treatment | Alcohol use during pregnancy (excluded) |
| `wpre` | Confounder | Pre-pregnancy weight (lbs) |
| `education` | Confounder | Years of education |
| `wgain`, `gestation` | Post-treatment | Weight gain, gestation weeks (excluded) |

```{r setup-data}
# ── Packages ──────────────────────────────────────────────────────────────────
library(SuperLearner)  # install.packages("SuperLearner")
library(lmtest)        # install.packages("lmtest")
library(sandwich)      # install.packages("sandwich")
library(dplyr)
library(ggplot2)
library(tibble)

# ── Reproducibility ───────────────────────────────────────────────────────────
set.seed(123)

# ── Load data ─────────────────────────────────────────────────────────────────
# Update the path below to point to your local copy of KingCounty2001.data
file_path <- "KingCounty2001.data"

df <- read.table(file_path)
colnames(df) <- c("gender", "plural", "age", "race", "parity", "married",
                  "bwt", "smokeN", "drinkN", "firstep", "welfare",
                  "smoker", "drinker", "wpre", "wgain", "education", "gestation")

# ── Initial inspection ────────────────────────────────────────────────────────
glimpse(df)
```

------------------------------------------------------------------------

## 1. Data Preparation

We construct the binary outcome (low birth weight), recode categorical
variables, and select the confounders, **excluding post-treatment
variables** (`smoker`, `drinker`, `wgain`, `gestation`) to avoid
mediation and collider bias.

```{r data-prep}
N <- nrow(df)

# Binary outcome: low birth weight (< 2500 g)
df$low_bwt <- as.integer(df$bwt < 2500)

# Recode character variables to numeric
df$gender  <- as.integer(df$gender  == "M")   # 1 = male
df$smoker  <- as.integer(df$smoker  == "Y")   # 1 = smoker
df$drinker <- as.integer(df$drinker == "Y")   # 1 = drinker
df$married <- as.integer(df$married == "Y")   # 1 = married

# Analysis datasets ────────────────────────────────────────────────────────────
# Q1 & Q2: exposure = First Steps (binary), outcome = low_bwt
Y <- df$low_bwt
A <- df$firstep
L <- df[, c("gender", "age", "race", "parity", "married",
            "smoker", "drinker", "wpre", "education")]

# Q3: exposure = maternal age (continuous), outcome = low_bwt
# Exclude parity & firstep: potential mediators / not confounders of age→bwt
y <- df$low_bwt
a <- df$age
l <- df[, c("gender", "married", "smoker", "drinker", "wpre", "education")]
```

------------------------------------------------------------------------

## 2. Exploratory Data Analysis

### 2.1 Outcome and exposure distributions

```{r eda-outcome-exposure, fig.width=10, fig.height=4}
p1 <- df %>%
  mutate(firstep = factor(firstep, labels = c("Non-participant", "Participant"))) %>%
  ggplot(aes(x = firstep, fill = firstep)) +
  geom_bar(alpha = 0.8, width = 0.4) +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  theme_minimal() +
  labs(x = NULL, y = "Count", title = "First Steps participation") +
  theme(legend.position = "none")

p2 <- df %>%
  mutate(low_bwt = factor(low_bwt, labels = c("Normal weight", "Low birth weight"))) %>%
  ggplot(aes(x = low_bwt, fill = low_bwt)) +
  geom_bar(alpha = 0.8, width = 0.4) +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  theme_minimal() +
  labs(x = NULL, y = "Count", title = "Low birth weight outcome") +
  theme(legend.position = "none")

library(patchwork)
p1 + p2
```

### 2.2 Maternal age distribution

```{r eda-age, fig.width=7, fig.height=4}
ggplot(df, aes(x = age)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30,
                 fill = "steelblue", alpha = 0.7) +
  geom_density(color = "navy", linewidth = 0.8) +
  theme_minimal() +
  labs(x = "Maternal age (years)", y = "Density",
       title = "Distribution of maternal age")
```

### 2.3 Outcome by First Steps participation

```{r eda-lbw-by-firstep, fig.width=7, fig.height=4}
df %>%
  mutate(firstep = factor(firstep, labels = c("Non-participant", "Participant")),
         low_bwt = factor(low_bwt, labels = c("Normal", "Low BW"))) %>%
  count(firstep, low_bwt) %>%
  group_by(firstep) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = firstep, y = prop, fill = low_bwt)) +
  geom_col(position = "dodge", alpha = 0.85, width = 0.5) +
  geom_text(aes(label = scales::percent(prop, accuracy = 0.1)),
            position = position_dodge(width = 0.5), vjust = -0.4, size = 3.5) +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  theme_minimal() +
  labs(x = NULL, y = "Proportion", fill = NULL,
       title = "Low birth weight rate by First Steps participation")
```

------------------------------------------------------------------------

## 3. Super Learner Setup

We use `SuperLearner` to estimate the nuisance functions
$\hat{E}(Y \mid L)$ (outcome model, $\hat{q}$) and $\hat{E}(A \mid L)$
(exposure model / propensity score, $\hat{p}$).

We start with a **small library** for speed, and extend it in Question
4.

```{r sl-setup}
SL.library <- c("SL.glm", "SL.ranger")
```

> **Note on cross-fitting:** Without cross-fitting, nuisance functions
> are fit on the same data used for the final estimator, which can
> introduce overfitting bias. Cross-fitting (i.e. fitting on one fold
> and predicting on another) removes this bias. We implement 2-fold
> cross-fitting in Question 4.

------------------------------------------------------------------------

## 4. Question 1 — AL Linear Regression: First Steps → Low Birth Weight

### 4.1 Relevant confounders

Post-treatment variables (`smoker`, `drinker`, `wgain`, `gestation`) are
**excluded** because adjusting for them would induce mediation or
collider bias. Variables included as confounders are those measured
*before* treatment that may jointly influence First Steps participation
and birth weight: maternal age, race, parity, marital status,
pre-pregnancy weight, and education.

Possible **missing confounders** include household income (only proxied
by education) and health insurance status, which affects both
eligibility for First Steps and access to prenatal care.

### 4.2 Estimating nuisance functions (no cross-fitting)

```{r q1-nuisance}
# Outcome model: E(Y | L)
q.model <- SuperLearner(Y = Y, X = L,
                        SL.library = SL.library,
                        family = "binomial",
                        cvControl = list(V = 5))
q <- q.model$SL.predict

# Exposure model: E(A | L) = P(A=1 | L)
p.model <- SuperLearner(Y = A, X = L,
                        SL.library = SL.library,
                        family = "binomial",
                        cvControl = list(V = 5))
p <- p.model$SL.predict
```

### 4.3 Assumption-lean estimator via `lm` (sandwich SE)

```{r q1-lm}
# Regress (Y - q_hat) onto (A - p_hat), no intercept
model_q1 <- lm(I(Y - q) ~ -1 + I(A - p))
lm_res_q1 <- coeftest(model_q1, vcov = sandwich)

beta_lm <- lm_res_q1[1, 1]
se_lm   <- lm_res_q1[1, 2]
```

### 4.4 Assumption-lean estimator via EIF (by hand)

```{r q1-eif}
# Point estimate from the efficient influence function
beta_eif <- mean((A - p) * (Y - q)) / mean((A - p)^2)

# Standard error from the EIF
se_eif <- (1 / sqrt(N)) *
  sd((A - p) * (Y - q - beta_eif * (A - p)) / mean((A - p)^2))
```

### 4.5 Results

```{r q1-results}
results_q1 <- tibble(
  Method    = c("Linear Model", "Eff. Influence Function"),
  Estimate  = c(beta_lm,  beta_eif),
  StdError  = c(se_lm,    se_eif),
  CI_Lower  = c(beta_lm  - 1.96 * se_lm,  beta_eif  - 1.96 * se_eif),
  CI_Upper  = c(beta_lm  + 1.96 * se_lm,  beta_eif  + 1.96 * se_eif)
)

knitr::kable(results_q1, digits = 4,
             caption = "AL linear regression: First Steps on low birth weight")
```

> **Interpretation (assuming constant effect):** The estimate
> $\hat\beta$ represents the overlap-weighted average risk difference
> $E\{w(L)\beta(L)\}/E\{w(L)\}$. If $\beta(L) = \beta$ (constant), this
> equals the population average treatment effect. Participation in First
> Steps is estimated to **reduce** the probability of low birth weight
> by approximately 0.39 percentage points; however, this estimate is
> **not statistically significant** (95% CI includes 0).
>
> **If** $\beta(L)$ is not constant: The effect of First Steps likely
> varies by individual characteristics (e.g. younger or less-educated
> mothers may benefit more). The estimand is then a weighted average
> that gives higher weight to covariate strata with greater treatment
> variability—those where we learn most from the data.
>
> **Advantage over standard linear regression:** If the outcome model is
> misspecified, the standard OLS coefficient no longer captures the
> association of interest and may even be non-zero when
> $A \perp\!\!\!\perp Y \mid L$. The assumption-lean estimator remains
> interpretable regardless of misspecification, and valid data-adaptive
> inference is obtained through the EIF.

### 4.6 Comparison with AIPW and TMLE (optional)

```{r q1-aipw-tmle, eval=FALSE}
# AIPW (npcausal package)
library(npcausal)
ate.aipw <- ate(y = Y, a = A, x = L, nsplits = 5, sl.lib = SL.library)

# TMLE
library(tmle)
ate.tmle <- tmle(Y = Y, A = A, W = L,
                 Q.SL.library = SL.library,
                 g.SL.library = SL.library,
                 family = "binomial", V.Q = 5, V.g = 5)
```

> Note: `npcausal` assumes a Gaussian outcome by default. Results from
> AIPW and TMLE target the average treatment effect (ATE) under a causal
> interpretation, whereas the AL estimator targets the overlap-weighted
> association.

------------------------------------------------------------------------

## 5. Question 2 — Weighted Baseline Characteristics

Assumption-lean regression implicitly weights participants by
$w(L) = P(A=1 \mid L)\,P(A=0 \mid L)$. It is therefore informative to
report **weighted summary statistics** to characterize the effective
study population.

### 5.1 Why not plug in directly?

The naive plug-in estimator
$$\hat\theta_{\text{plug-in}} = \frac{\sum_i \hat{p}_i(1-\hat{p}_i)\,X_i}{\sum_i \hat{p}_i(1-\hat{p}_i)}$$
suffers **plug-in bias**: machine learning estimates $\hat p_i$ are
regularized (biased toward zero shrinkage) and the product
$\hat{p}_i(1-\hat{p}_i)$ is a non-linear function of $p_i$, so the bias
does not cancel in the ratio.

### 5.2 EIF-based weighted summary

The efficient influence function of
$\theta = E\{P(A=1|L)P(A=0|L)\,X\}/E\{P(A=1|L)P(A=0|L)\}$ is:

$$
\mathrm{IF}(\theta) = -\frac{\bigl[P(A=1|L)P(A=0|L) + \{1 - 2P(A=1|L)\}\{A - P(A=1|L)\}\bigr](X - \theta)}{E\{P(A=1|L)P(A=0|L)\}}.
$$

It can be shown that $P(A=1|L)P(A=0|L) + \{1-2p\}(A-p) = (A-p)^2$, so
the debiased estimator simplifies to:

$$
\hat\theta = \frac{\sum_i (A_i - \hat p_i)^2\,X_i}{\sum_i (A_i - \hat p_i)^2},
\qquad
\widehat{\mathrm{SE}}(\hat\theta) = \frac{1}{\sqrt{n}}\,\mathrm{sd}\!\left(\frac{(A_i-\hat p_i)^2(X_i-\hat\theta)}{E[p(1-p)]}\right).
$$

```{r q2-weighted-summary}
# Recode non-numeric variables used in L
L_num <- L
# (already recoded: gender=0/1, smoker=0/1, drinker=0/1, married=0/1)
# race is non-numeric → skipped automatically below

weighted_summary <- do.call(rbind, lapply(names(L_num), function(varname) {
  X <- L_num[[varname]]
  if (!is.numeric(X)) return(NULL)

  w_mean <- mean((A - p)^2 * X) / mean((A - p)^2)
  eif    <- (A - p)^2 * (X - w_mean) / mean(p * (1 - p))
  w_se   <- sd(eif) / sqrt(N)

  data.frame(
    variable    = varname,
    w_mean      = round(w_mean, 3),
    w_se        = round(w_se,   3),
    w_CI_lower  = round(w_mean - 1.96 * w_se, 3),
    w_CI_upper  = round(w_mean + 1.96 * w_se, 3)
  )
}))

knitr::kable(weighted_summary,
             col.names = c("Variable", "Weighted mean", "SE", "CI lower", "CI upper"),
             caption   = "Weighted baseline characteristics (overlap weights)")
```

### 5.3 Classic (unweighted) summary for comparison

```{r q2-classic-summary}
classic_summary <- do.call(rbind, lapply(names(L_num), function(varname) {
  X <- L_num[[varname]]
  if (!is.numeric(X)) return(NULL)

  c_mean <- mean(X)
  c_se   <- sd(X) / sqrt(N)

  data.frame(
    variable    = varname,
    c_mean      = round(c_mean, 3),
    c_se        = round(c_se,   3),
    c_CI_lower  = round(c_mean - 1.96 * c_se, 3),
    c_CI_upper  = round(c_mean + 1.96 * c_se, 3)
  )
}))

knitr::kable(classic_summary,
             col.names = c("Variable", "Classic mean", "SE", "CI lower", "CI upper"),
             caption   = "Classic (unweighted) baseline characteristics")
```

> **Interpretation:** Differences between weighted and unweighted means
> reveal which subpopulation the AL estimator effectively targets.
> Participants with intermediate propensity scores (neither near 0 nor
> 1) receive the highest overlap weight, since they carry the most
> information about the treatment effect.

------------------------------------------------------------------------

## 6. Question 3 — AL Linear Regression: Maternal Age → Low Birth Weight

Here the **exposure is continuous** (maternal age). The estimand is:

$$
\beta = \frac{E\{\mathrm{Cov}(A, Y \mid L)\}}{E\{\mathrm{Var}(A \mid L)\}},
$$

interpreted as the average change in risk of low birth weight associated
with a 1-year increase in maternal age, after adjusting for confounders
$L$. This corresponds to a **shift intervention**: if everyone's age
were shifted by $a$ years, the expected outcome would change by
$\approx \beta \cdot a$.

We exclude `parity` and `firstep` from the confounder set since they may
mediate the age effect.

```{r q3-nuisance}
# Outcome model: E(Y | l)  [binary outcome]
q2.model <- SuperLearner(Y = y, X = l,
                         SL.library = SL.library,
                         family = "binomial",
                         cvControl = list(V = 5))
q2 <- q2.model$SL.predict

# Exposure model: E(A | l) for continuous A
p2.model <- SuperLearner(Y = a, X = l,
                         SL.library = SL.library,
                         family = "gaussian",
                         cvControl = list(V = 5))
p2 <- p2.model$SL.predict
```

```{r q3-estimates}
# LM with sandwich SE
model_q3  <- lm(I(y - q2) ~ -1 + I(a - p2))
lm_res_q3 <- coeftest(model_q3, vcov = sandwich)

beta_lm3 <- lm_res_q3[1, 1]
se_lm3   <- lm_res_q3[1, 2]

# EIF
beta_eif3 <- mean((a - p2) * (y - q2)) / mean((a - p2)^2)
se_eif3   <- (1 / sqrt(N)) *
  sd((a - p2) * (y - q2 - beta_eif3 * (a - p2)) / mean((a - p2)^2))

results_q3 <- tibble(
  Method   = c("Linear Model", "Eff. Influence Function"),
  Estimate = c(beta_lm3, beta_eif3),
  StdError = c(se_lm3,   se_eif3),
  CI_Lower = c(beta_lm3 - 1.96 * se_lm3, beta_eif3 - 1.96 * se_eif3),
  CI_Upper = c(beta_lm3 + 1.96 * se_lm3, beta_eif3 + 1.96 * se_eif3)
)

knitr::kable(results_q3, digits = 5,
             caption = "AL linear regression: maternal age on low birth weight")
```

> **Interpretation (shift intervention):** A 1-year increase in maternal
> age is estimated to change the risk of low birth weight by
> $\hat\beta \approx 0.0017$ (roughly 0.17 percentage points). The 95%
> confidence interval includes 0, so the association is **not
> statistically significant**.

------------------------------------------------------------------------

## 7. Question 4 — Improved Estimates with Cross-Fitting and Richer Library

Cross-fitting removes the overfitting bias that arises when the same
observations are used to fit nuisance functions and to compute the
estimator. Here we use **2-fold cross-fitting** and add `SL.earth` to
the learner library.

```{r q4-setup}
SL.library2 <- c("SL.glm", "SL.randomForest", "SL.earth")

# Random split into two halves
I  <- sort(sample(1:N, N / 2))
IC <- setdiff(1:N, I)
```

### 7.1 First Steps program (binary exposure)

```{r q4-firstep, cache=TRUE}
# Fit on IC, predict on I
q_IC.mod <- SuperLearner(Y = Y[IC], X = L[IC, ], SL.library = SL.library2,
                         family = "binomial", cvControl = list(V = 5))
p_IC.mod <- SuperLearner(Y = A[IC], X = L[IC, ], SL.library = SL.library2,
                         family = "binomial", cvControl = list(V = 5))

# Fit on I, predict on IC
q_I.mod  <- SuperLearner(Y = Y[I],  X = L[I, ],  SL.library = SL.library2,
                         family = "binomial", cvControl = list(V = 5))
p_I.mod  <- SuperLearner(Y = A[I],  X = L[I, ],  SL.library = SL.library2,
                         family = "binomial", cvControl = list(V = 5))

# Cross-fitted predictions
q_cf <- numeric(N)
p_cf <- numeric(N)
q_cf[I]  <- predict(q_IC.mod, newdata = L[I, ],  onlySL = TRUE)$pred
q_cf[IC] <- predict(q_I.mod,  newdata = L[IC, ], onlySL = TRUE)$pred
p_cf[I]  <- predict(p_IC.mod, newdata = L[I, ],  onlySL = TRUE)$pred
p_cf[IC] <- predict(p_I.mod,  newdata = L[IC, ], onlySL = TRUE)$pred

# Estimates
model_q4a  <- lm(I(Y - q_cf) ~ -1 + I(A - p_cf))
lm_res_q4a <- coeftest(model_q4a, vcov = sandwich)

beta_lm_cf <- lm_res_q4a[1, 1]
se_lm_cf   <- lm_res_q4a[1, 2]

beta_eif_cf <- mean((A - p_cf) * (Y - q_cf)) / mean((A - p_cf)^2)
se_eif_cf   <- (1 / sqrt(N)) *
  sd((A - p_cf) * (Y - q_cf - beta_eif_cf * (A - p_cf)) / mean((A - p_cf)^2))

results_q4a <- tibble(
  Method   = c("Linear Model", "Eff. Influence Function"),
  Estimate = c(beta_lm_cf, beta_eif_cf),
  StdError = c(se_lm_cf,   se_eif_cf),
  CI_Lower = c(beta_lm_cf - 1.96 * se_lm_cf, beta_eif_cf - 1.96 * se_eif_cf),
  CI_Upper = c(beta_lm_cf + 1.96 * se_lm_cf, beta_eif_cf + 1.96 * se_eif_cf)
)

knitr::kable(results_q4a, digits = 4,
             caption = "Q4 – Cross-fitted AL: First Steps on low birth weight")
```

### 7.2 Maternal age (continuous exposure)

```{r q4-age, cache=TRUE}
q4_IC.mod <- SuperLearner(Y = y[IC], X = l[IC, ], SL.library = SL.library2,
                          family = "binomial", cvControl = list(V = 5))
p4_IC.mod <- SuperLearner(Y = a[IC], X = l[IC, ], SL.library = SL.library2,
                          family = "gaussian", cvControl = list(V = 5))

q4_I.mod  <- SuperLearner(Y = y[I],  X = l[I, ],  SL.library = SL.library2,
                          family = "binomial", cvControl = list(V = 5))
p4_I.mod  <- SuperLearner(Y = a[I],  X = l[I, ],  SL.library = SL.library2,
                          family = "gaussian", cvControl = list(V = 5))

q4_cf <- numeric(N)
p4_cf <- numeric(N)
q4_cf[I]  <- predict(q4_IC.mod, newdata = l[I, ],  onlySL = TRUE)$pred
q4_cf[IC] <- predict(q4_I.mod,  newdata = l[IC, ], onlySL = TRUE)$pred
p4_cf[I]  <- predict(p4_IC.mod, newdata = l[I, ],  onlySL = TRUE)$pred
p4_cf[IC] <- predict(p4_I.mod,  newdata = l[IC, ], onlySL = TRUE)$pred

model_q4b  <- lm(I(y - q4_cf) ~ -1 + I(a - p4_cf))
lm_res_q4b <- coeftest(model_q4b, vcov = sandwich)

beta_lm4b <- lm_res_q4b[1, 1]
se_lm4b   <- lm_res_q4b[1, 2]

beta_eif4b <- mean((a - p4_cf) * (y - q4_cf)) / mean((a - p4_cf)^2)
se_eif4b   <- (1 / sqrt(N)) *
  sd((a - p4_cf) * (y - q4_cf - beta_eif4b * (a - p4_cf)) / mean((a - p4_cf)^2))

results_q4b <- tibble(
  Method   = c("Linear Model", "Eff. Influence Function"),
  Estimate = c(beta_lm4b, beta_eif4b),
  StdError = c(se_lm4b,   se_eif4b),
  CI_Lower = c(beta_lm4b - 1.96 * se_lm4b, beta_eif4b - 1.96 * se_eif4b),
  CI_Upper = c(beta_lm4b + 1.96 * se_lm4b, beta_eif4b + 1.96 * se_eif4b)
)

knitr::kable(results_q4b, digits = 5,
             caption = "Q4 – Cross-fitted AL: maternal age on low birth weight")
```

------------------------------------------------------------------------

## 8. Question 5 — Risk Ratio Scale

We use **assumption-lean log-linear regression**:

$$
\log E(Y^a \mid L) = \alpha(L) + \beta a,
$$

so that
$\beta = \log\bigl(E[Y^1 \mid L] / E[Y^0 \mid L]\bigr) = \log(\mathrm{RR})$
and $\mathrm{RR} = e^\beta$.

The debiased estimator regresses the pseudo-outcome

$$
\widetilde{Y}^{\log} = \log(\hat{Y}) - \hat{q}^{\log} + \frac{Y - \hat{Y}}{\hat{Y}}
$$

onto $(A - \hat{p})$ with no intercept (sandwich SE), where
$\hat{q}^{\log} = \hat{E}[\log(\hat{Y}) \mid L]$.

### 8.1 First Steps program

```{r q5-rr-firstep, cache=TRUE}
# Step 1: fit E(Y | A, L) using both A and L
AL <- cbind(A = A, L)
Y_hat.mod <- SuperLearner(Y = Y, X = AL,
                          SL.library = SL.library,
                          family = "binomial",
                          cvControl = list(V = 5))
Y_hat <- Y_hat.mod$SL.predict

# Step 2: fit E[log(Y_hat) | L]
log_Y_hat <- log(Y_hat)
q_log.mod <- SuperLearner(Y = log_Y_hat, X = L,
                          SL.library = SL.library,
                          family = "gaussian",
                          cvControl = list(V = 5))
q_log <- q_log.mod$SL.predict

# Step 3: construct pseudo-outcome and regress
pseudo_log <- log_Y_hat - q_log + (Y - Y_hat) / Y_hat
log_mod    <- lm(I(pseudo_log) ~ -1 + I(A - p))
log_res    <- coeftest(log_mod, vcov = sandwich)

log_beta <- log_res[1, 1]
log_se   <- log_res[1, 2]

RR_results <- tibble(
  Method   = "Risk Ratio",
  Estimate = exp(log_beta),
  StdError = log_se,
  CI_Lower = exp(log_beta - 1.96 * log_se),
  CI_Upper = exp(log_beta + 1.96 * log_se)
)

knitr::kable(RR_results, digits = 3,
             caption = "Risk ratio – First Steps program on low birth weight")
```

### 8.2 Maternal age

```{r q5-rr-age, cache=TRUE}
al <- cbind(a = a, l)
y_hat.mod <- SuperLearner(Y = y, X = al,
                          SL.library = SL.library,
                          family = "binomial",
                          cvControl = list(V = 5))
y_hat <- y_hat.mod$SL.predict

log_y_hat  <- log(y_hat)
q2_log.mod <- SuperLearner(Y = log_y_hat, X = l,
                           SL.library = SL.library,
                           family = "gaussian",
                           cvControl = list(V = 5))
q2_log <- q2_log.mod$SL.predict

pseudo_log2 <- log_y_hat - q2_log + (y - y_hat) / y_hat
log_mod2    <- lm(I(pseudo_log2) ~ -1 + I(a - p2))
log_res2    <- coeftest(log_mod2, vcov = sandwich)

log_beta2 <- log_res2[1, 1]
log_se2   <- log_res2[1, 2]

RR_results2 <- tibble(
  Method   = "Risk Ratio Age",
  Estimate = exp(log_beta2),
  StdError = log_se2,
  CI_Lower = exp(log_beta2 - 1.96 * log_se2),
  CI_Upper = exp(log_beta2 + 1.96 * log_se2)
)

knitr::kable(RR_results2, digits = 4,
             caption = "Risk ratio – maternal age on low birth weight")
```

> **Interpretation:** A risk ratio of 1 (or log-RR of 0) corresponds to
> no effect. The First Steps risk ratio and the per-year age risk ratio
> are both close to 1 and not statistically significant.

------------------------------------------------------------------------

## 9. Question 6 — Odds Ratio Scale

We use **assumption-lean logistic regression**:

$$
\mathrm{logit}\,E(Y^a \mid L) = \alpha(L) + \beta a,
$$

so that $\beta = \log(\mathrm{OR})$ and $\mathrm{OR} = e^\beta$.

The pseudo-outcome is now:

$$
\widetilde{Y}^{\mathrm{logit}} = \mathrm{logit}(\hat{Y}) - \hat{q}^{\mathrm{logit}} + \frac{Y - \hat{Y}}{\hat{Y}(1 - \hat{Y})},
$$

which is regressed onto $(A - \hat{p})$ with no intercept and a sandwich
SE.

### 9.1 First Steps program

```{r q6-or-firstep, cache=TRUE}
logit_Y_hat <- log(Y_hat / (1 - Y_hat))
q_logit.mod <- SuperLearner(Y = logit_Y_hat, X = L,
                            SL.library = SL.library,
                            family = "gaussian",
                            cvControl = list(V = 5))
q_logit <- q_logit.mod$SL.predict

pseudo_logit <- logit_Y_hat - q_logit + (Y - Y_hat) / (Y_hat * (1 - Y_hat))
logit_mod    <- lm(I(pseudo_logit) ~ -1 + I(A - p))
logit_res    <- coeftest(logit_mod, vcov = sandwich)

logit_beta <- logit_res[1, 1]
logit_se   <- logit_res[1, 2]

OR_results <- tibble(
  Method   = "Odds Ratio",
  Estimate = exp(logit_beta),
  StdError = logit_se,
  CI_Lower = exp(logit_beta - 1.96 * logit_se),
  CI_Upper = exp(logit_beta + 1.96 * logit_se)
)

knitr::kable(OR_results, digits = 3,
             caption = "Odds ratio – First Steps program on low birth weight")
```

### 9.2 Maternal age

```{r q6-or-age, cache=TRUE}
logit_y_hat  <- log(y_hat / (1 - y_hat))
q2_logit.mod <- SuperLearner(Y = logit_y_hat, X = l,
                             SL.library = SL.library,
                             family = "gaussian",
                             cvControl = list(V = 5))
q2_logit <- q2_logit.mod$SL.predict

pseudo_logit2 <- logit_y_hat - q2_logit + (y - y_hat) / (y_hat * (1 - y_hat))
logit_mod2    <- lm(I(pseudo_logit2) ~ -1 + I(a - p2))
logit_res2    <- coeftest(logit_mod2, vcov = sandwich)

logit_beta2 <- logit_res2[1, 1]
logit_se2   <- logit_res2[1, 2]

OR_results2 <- tibble(
  Method   = "Odds Ratio Age",
  Estimate = exp(logit_beta2),
  StdError = logit_se2,
  CI_Lower = exp(logit_beta2 - 1.96 * logit_se2),
  CI_Upper = exp(logit_beta2 + 1.96 * logit_se2)
)

knitr::kable(OR_results2, digits = 3,
             caption = "Odds ratio – maternal age on low birth weight")
```

### 9.3 Visualizing the odds-ratio transformation

An advantage of odds ratios is that they are bounded away from 0 (more
stable numerically) and arise naturally from logistic models. A
disadvantage is non-collapsibility: the marginal OR differs from the
conditional OR even without confounding.

To aid interpretation, we plot $E[Y \mid A=1, L]$ as a function of
$E[Y \mid A=0, L]$ implied by the estimated log-OR $\hat\beta$:

```{r q6-plot, fig.width=6, fig.height=5}
expit <- function(x) 1 / (1 + exp(-x))
logit <- function(p) log(p / (1 - p))

x_seq <- seq(0.001, 0.999, length.out = 1000)
y_seq <- expit(logit(x_seq) + logit_beta)

plot(x_seq, y_seq, type = "l", col = "steelblue", lwd = 2,
     xlab = expression(E(Y ~ "|" ~ A==0 ~ "," ~ L)),
     ylab = expression(E(Y ~ "|" ~ A==1 ~ "," ~ L)),
     main = expression("Implied" ~ E(Y ~ "|" ~ A==1 ~ "," ~ L) ~
                       "= expit(logit(" * hat(Y)[A==0] * ") + " * hat(beta) * ")"))
abline(0, 1, col = "gray50", lty = 2)
legend("topleft", legend = c("OR model", "Identity"),
       col = c("steelblue", "gray50"), lty = c(1, 2), bty = "n")
```

> When $\hat\beta$ is small (near 0), the curve nearly coincides with
> the identity line, confirming a near-null association. If $\hat\beta$
> were large, the S-shaped curve would highlight how the implied risk
> difference $E(Y|A=1,L) - E(Y|A=0,L)$ varies across the baseline risk
> spectrum—a key reason odds ratios are harder to interpret than risk
> differences.

------------------------------------------------------------------------

## 10. Summary of Results

```{r summary-table, fig.width=9, fig.height=5}
summary_all <- tibble(
  Exposure = rep(c("First Steps", "Maternal Age"), each = 3),
  Scale    = rep(c("Risk Difference", "Risk Ratio", "Odds Ratio"), times = 2),
  Estimate = c(beta_eif,        exp(log_beta),   exp(logit_beta),
               beta_eif3,       exp(log_beta2),  exp(logit_beta2)),
  SE       = c(se_eif,          log_se,          logit_se,
               se_eif3,         log_se2,         logit_se2),
  Null     = rep(c(0, 1, 1), 2)
)

summary_all <- summary_all %>%
  mutate(CI_Lower = if_else(Scale == "Risk Difference",
                            Estimate - 1.96 * SE, exp(log(Estimate) - 1.96 * SE)),
         CI_Upper = if_else(Scale == "Risk Difference",
                            Estimate + 1.96 * SE, exp(log(Estimate) + 1.96 * SE)))

knitr::kable(summary_all %>% select(-Null) %>% mutate(across(where(is.numeric), ~round(.x, 4))),
             caption = "Summary of all assumption-lean regression results")
```

```{r forest-plot, fig.width=9, fig.height=5}
summary_all %>%
  mutate(Label = paste(Exposure, "-", Scale),
         Label = factor(Label, levels = rev(unique(Label)))) %>%
  ggplot(aes(x = Estimate, y = Label, color = Exposure)) +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = CI_Lower, xmax = CI_Upper), height = 0.3, linewidth = 0.8) +
  geom_vline(aes(xintercept = Null), linetype = "dashed", color = "gray50") +
  facet_wrap(~ Scale, scales = "free_x", nrow = 1) +
  scale_color_manual(values = c("steelblue", "tomato")) +
  theme_minimal() +
  labs(x = "Estimate (95% CI)", y = NULL, color = "Exposure",
       title = "Assumption-Lean Regression — King County 2001",
       subtitle = "Risk difference, risk ratio, and odds ratio scales") +
  theme(legend.position = "bottom")
```

------------------------------------------------------------------------

## 11. Key Takeaways

The assumption-lean framework offers three main advantages over standard
regression:

1.  **Robustness to misspecification.** The estimand $\beta =
    E\{\mathrm{Cov}(A,Y|L)\}/E\{\mathrm{Var}(A|L)\}$ is well-defined
    regardless of whether the working model holds. Under
    misspecification, standard OLS may not even capture the sign of the
    association.

2.  **Valid data-adaptive inference.** By relying on the efficient
    influence function and cross-fitting, we obtain valid confidence
    intervals even when nuisance functions are estimated with machine
    learning or data-driven model selection.

3.  **Scale flexibility.** The same framework extends seamlessly to log
    and logit links, enabling risk ratio and odds ratio estimation
    without additional parametric assumptions.


