---
title: "Causal Machine Learning: Project"
author: "Frederick De Baene, Jonas Geysen, Antoon Goderis"
output: pdf
---

# Setup

```{r}
library(hdm)
library(SuperLearner)
library(tmle)
library(dplyr)
library(flextable)
library(ggplot2)
# library(rlearner)
library(grf)
```

```{r}
# Load data
data(pension)
```

```{r}
sl_lib <- list("SL.glm", "SL.step", "SL.glm.interaction", "SL.randomForest")
```

# Question 2

```{r}
# There are 9915 observations in total
n <- nrow(pension)

# The outcome of interest are the accumulated net financial assets
Y <- pension$net_tfa
  
# The exposure of interest is eligibility for a 401(k)
A <- pension$e401

# The covariates include the following
covariates <- c(
  "age",
  "inc",
  "fsize",
  "educ",
  "marr",
  "twoearn",
  "db",
  "pira",
  "hown"
)
X <- pension[, covariates, drop = FALSE]
```

We start by estimating the propensity scores non-parametrically using
SuperLearner. We apply 10-fold cross-fitting.

```{r}
set.seed(123)

# We apply 10-fold cross-fitting
K <- 5
fold_indices <- sample(rep(x = 1:K, times = ceiling(n / K))[1:n])

# Obtain propensity scores
mu_hat <- tibble(
  "mu" = rep(NA, times = n),
  "mu0" = rep(NA, times = n),
  "mu1" = rep(NA, times = n),
)
est_prop_scores <- rep(NA, n)

for (k in 1:K) {

  print(paste0("Iteration : ", k))
  train <- fold_indices != k
  test <- fold_indices == k
  
  if (K == 1)
    train <- test
  
  # Use SuperLearner to obtain propensity scores
  print("Estimating propensity scores ...")
  model <- SuperLearner(
    A[train],
    X[train,],
    newX = X[test,],
    family = binomial(link = "logit"),
    SL.library = sl_lib
  )
  est_prop_scores[test] <- model$SL.predict
  
  # Use SuperLearner to obtain estimated outcomes
  print("Estimating outcomes in non-exposed ...")
  model0 <- SuperLearner(
    Y[(A == 0) & train],
    X[(A == 0) & train,],
    newX = X[test,],
    family = "gaussian",
    SL.library = sl_lib
  )
  mu_hat[test, "mu0"] <- model0$SL.predict
  
  print("Estimating outcomes in exposed ...")
  model1 <- SuperLearner(
    Y[(A == 1) & train],
    X[(A == 1) & train,],
    newX = X[test,],
    family = "gaussian",
    SL.library = sl_lib
  )
  mu_hat[test, "mu1"] <- model1$SL.predict
  
}

mu_hat <- dplyr::mutate(mu_hat, mu = A*mu1 + (1 - A)*mu0)
```

```{r}
saveRDS(est_prop_scores, file = "est_prop_scores_ATE_K_5.rds")
saveRDS(mu_hat, file = "mu_hat_ATE_K_5.rds")
```

```{r}
est_prop_scores <- readRDS("est_prop_scores_ATE_K_5.rds")
mu_hat <- readRDS("mu_hat_ATE_K_5.rds")
```

```{r}
# Obtain the estimated probability of exposure and non-exposure given the 
# confounders
G1_aipw <- est_prop_scores
G0_aipw <- 1 - est_prop_scores

# Obtain the estimated outcome given exposure and non-exposure
Q_aipw <- mu_hat$mu
Q1_aipw <- mu_hat$mu1
Q0_aipw <- mu_hat$mu0
```

We calculate the estimate.

```{r}
ate_aipw <- mean(((A / G1_aipw) - ((1 - A) / G0_aipw)) * (Y - Q_aipw) + Q1_aipw - Q0_aipw)
ate_aipw
```

To calculate the standard error, we apply the efficient influence curve to
every observation. The standard error is calculated using the standard deviation
of these values.

```{r}
ate_IF <- (A / G1_aipw - (1 - A) / G0_aipw) * (Y - Q_aipw) + Q1_aipw - Q0_aipw - ate_aipw
ate_aipw_se <- sd(ate_IF) / sqrt(n)
ate_aipw_se
```

We can also construct a confidence interval:

```{r}
ll <- ate_aipw - 1.96 * ate_aipw_se; ll
ul <- ate_aipw + 1.96 * ate_aipw_se; ul
```

```{r}
att_aipw_num <- sum(((A / mean(A)) * (Y - Q0_aipw)) - ((1 - A) / (G0_aipw)) * ((G1_aipw) / (mean(A))) * (Y - Q0_aipw))
att_aipw_den <- sum(A / mean(A))
att_aipw <- att_aipw_num / att_aipw_den
att_aipw
```

```{r}
att_IF <- (A / mean(A)) * (Y - Q0_aipw - att_aipw) - ((1 - A) / (G0_aipw)) * ((G1_aipw) / (mean(A))) * (Y - Q0_aipw)
att_aipw_se <- sd(att_IF) / sqrt(n)
att_aipw_se
```

```{r}
ll <- att_aipw - 1.96 * att_aipw_se; ll
ul <- att_aipw + 1.96 * att_aipw_se; ul
```

We also check the positivity assumption.

```{r}
# Check positivity assumption (c)
Apihat <- as.data.frame(cbind(A, est_prop_scores))

pi_A1 <- Apihat$est_prop_scores[Apihat$A == 1]
pi_A0 <- Apihat$est_prop_scores[Apihat$A == 0]

pi_A1_inverse <- (1/Apihat$est_prop_scores[Apihat$A == 1])
pi_A0_inverse <- (1/Apihat$est_prop_scores[Apihat$A == 0])
```

```{r}
png(filename="/home/rstudio/Project/imgs/prop_scores_dens.png")
plot(
  density(pi_A1, na.rm = TRUE),
  col = "red",
  lwd = 2, 
  main = "Density for Propensity Scores",
  xlab = "Propensity score",
  ylim = c(0, max(density(pi_A1)$y, density(pi_A0)$y))
)
lines(density(pi_A0, na.rm = TRUE), col = "blue", lwd = 2)
legend(
  "topright",
  legend = c("A = 1", "A = 0"),
  col = c("red", "blue"),
  lwd = 2,
)
dev.off()
```

```{r}
# Create a box plot with the propensity scores
png(filename="/home/rstudio/Project/imgs/prop_scores_boxp.png")
boxplot(
  pi_A1_inverse,
  pi_A0_inverse, 
  names = c("A = 1", "A = 0"), 
  col = c("red", "blue"), 
  main = "Distribution of Weights (Inverse of Propensity Scores)", 
  ylab = "1/pi_hat_nonpar_bounded"
)
dev.off()
```

```{r}
g <- est_prop_scores
Q <- mu_hat
names(Q) <- c("Q", "Q0", "Q1")
```

```{r}
N <- n
H1 <- A / g
H0 <- (1 - A) / (1 - g)

# Estimate parameters delta in the fluctuation model and target the predictions
Q.tmle <- as_tibble(matrix(0, nrow = N, ncol = 2,
                           dimnames = list(c(), c("Q0.tmle", "Q1.tmle"))))

delta1.tmle <- coef(lm(Y ~ -1 + offset(Q$Q1) + H1))
Q.tmle["Q1.tmle"] <- Q$Q1 + delta1.tmle * H1

delta0.tmle <- coef(lm(Y ~ -1 + offset(Q$Q0) + H0))
Q.tmle["Q0.tmle"] <- Q$Q0 + delta0.tmle * H0

# Compute E[Y^1] and E[Y^0]
EY1.tmle <- mean(Q.tmle$Q1.tmle)
EY0.tmle <- mean(Q.tmle$Q0.tmle)
ATE.tmle <- EY1.tmle - EY0.tmle

# Compute standard errors of E[Y^1] and E[Y^0] using the efficient influence curves
EIC.EY1.tmle <- (A / g) * (Y - Q.tmle$Q1.tmle) + Q.tmle$Q1.tmle - EY1.tmle
EIC.EY0.tmle <- ((1 - A) / (1 - g)) * (Y - Q.tmle$Q0.tmle) + Q.tmle$Q0.tmle - EY0.tmle

se.EY1.tmle <- (1 / sqrt(N)) * sd(EIC.EY1.tmle)
se.EY0.tmle <- (1 / sqrt(N)) * sd(EIC.EY0.tmle)
se.ATE.tmle <- (1 / sqrt(N)) * sd(EIC.EY1.tmle - EIC.EY0.tmle)

# Compute the confidence intervals for E[Y^1] and E[Y^0]
EY1.tmle.CI.low <- EY1.tmle - 1.96 * se.EY1.tmle
EY1.tmle.CI.high <- EY1.tmle + 1.96 * se.EY1.tmle

EY0.tmle.CI.low <- EY0.tmle - 1.96 * se.EY0.tmle
EY0.tmle.CI.high <- EY0.tmle + 1.96 * se.EY0.tmle

ATE.tmle.CI.low <- ATE.tmle - 1.96 * se.ATE.tmle
ATE.tmle.CI.high <- ATE.tmle + 1.96 * se.ATE.tmle

# Print the results
print(ATE.tmle)
print(ATE.tmle.CI.low)
print(ATE.tmle.CI.high)
```

For the ATT.

```{r}
P_A1 <- mean(A)
g <- G1_aipw
g <- pmin(g, 0.99)
g <- pmax(g, 0.01)

# Construct clever covariate H
H <- ifelse(A == 1, 1 / P_A1, -g / (P_A1 * (1 - g)))

# Fit fluctuation model (update Q0 to Q0_star)
epsilon_model <- glm(Y ~ -1 + H + offset(Q0_aipw), family = gaussian)
epsilon <- coef(epsilon_model)["H"]
Q0_tmle <- Q0_aipw + epsilon * H

# Compute ATT
theta_hat <- mean( (Y[A == 1] - Q0_tmle[A == 1]) )

# Influence function and standard error
IF_part1 <- (A / P_A1) * (Y - Q0_tmle - theta_hat)
IF_part2 <- ((1 - A) * g / (P_A1 * (1 - g))) * (Y - Q0_tmle)
IF <- IF_part1 - IF_part2
se <- sqrt(mean(IF^2)) / sqrt(n)
```

```{r}
theta_hat
```

```{r}
# Confidence interval
ci <- theta_hat + c(-1.96, 1.96) * se
ci
```

Calculate the ATT without covariate adjustment:

```{r}
Y_0 <- Y[A == 0]
Y_1 <- Y[A == 1]

ATT_wo_cov_adj <- mean(Y_1) - mean(Y_0)
ATT_wo_cov_adj
```

# Question 3

We estimate $\theta$ as follows:

```{r}
est_theta <- mean((A - est_prop_scores) * (Y - mu_hat$mu))
est_theta
```

We calculate the efficient influence curve for every observation:

```{r}
IF_theta <- (A - est_prop_scores) * (Y - mu_hat$mu) - est_theta
est_theta_se <- sd(IF_theta) / sqrt(n)
est_theta_se
```

To obtain a p-value:

```{r}
(1 - pnorm(q = est_theta, mean = 0, sd = est_theta_se)) * 2
```

# Question 4

```{r}
set.seed(123)

################################################################################
# 1. Create K folds
n <- nrow(pension)
folds <- sample(rep(1:5, length.out = n))

# 2. Initialize a matrix for CATEs
cate_estimates <- rep(NA, nrow(X))

# 3. Loop over folds
for (i in 1:5) {
  train_idx <- which(folds != i)
  test_idx  <- which(folds == i)
  
  # Fit causal forest on training data
  cf_model <- causal_forest(X[train_idx, ], 
                            Y[train_idx], 
                            A[train_idx],
                            tune.parameters = c("sample.fraction", 
                                                "mtry", 
                                                "min.node.size"))
  
  # Predict on the test fold
  cate_estimates[test_idx] <- predict(cf_model, X[test_idx, ])$predictions
}

################################################################################
# histogram
hist(cate_estimates,
     breaks = 30,
     main = "Histogram of Estimated CATEs",
     xlab = "Estimated Treatment Effect",
     col = "lightblue",
     border = "black")
mean(cate_estimates)
summary(cate_estimates)

################################################################################
results <- data.frame(tau = cate_estimates, age = X$age, inc = X$inc, educ = X$educ)

# By age
ggplot(results, aes(x = age, y = tau)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess") +
  labs(title = "CATE vs Age", x = "Age", y = "Estimated Treatment Effect") +
  theme_classic()

# By income
ggplot(results, aes(x = inc, y = tau)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess") +
  labs(title = "CATE vs Income", x = "Income", y = "Estimated Treatment Effect") +
  theme_classic()

# By education
ggplot(results, aes(x = educ, y = tau)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess") +
  labs(title = "CATE vs Education", x = "Years of Education", y = "Estimated Treatment Effect") +
  theme_classic()
```

# Question 5

```{r}
df <- pension

# Full covariate set (controls) used for nuisance estimation:
#X_full <- df[, c("age", "inc", "educ", "fsize", "marr", "twoearn", "db", "ira", "hown")]
X_full <- df[, c("age", "inc", "educ", "fsize", "marr", "twoearn", "db", "pira", "hown")]

# Reduced set for prediction (e.g., for visualizing heterogeneity or estimating the counterfactual outcome)
X_reduced <- df[, c("age", "inc", "educ")]

n <- nrow(X_full)
K <- 5

sl_lib <- c("SL.glm", "SL.randomForest", "SL.xgboost")

# Create folds for cross-fitting
folds <- sample(rep(1:K, length.out = n))
Y1_hat <- rep(NA, n)
p_hat <- rep(NA, n)

# STEP 2: Estimate Nuisance Functions with Cross-Fitting Using Full Covariates
for (k in 1:K) {
  train_idx <- which(folds != k)
  test_idx  <- which(folds == k)
  
  # Outcome model for treated subjects (to estimate Y1 for those with A == 1)
  mu1_fit <- SuperLearner(
    Y = Y[train_idx][A[train_idx] == 1],
    X = X_full[train_idx, ][A[train_idx] == 1, ],
    newX = X_full[test_idx, ],
    SL.library = sl_lib,
    family = gaussian()
  )
  Y1_hat[test_idx] <- mu1_fit$SL.predict
  
  # Propensity score estimation using full covariates
  ps_fit <- SuperLearner(
    Y = A[train_idx],
    X = X_full[train_idx, ],
    newX = X_full[test_idx, ],
    SL.library = sl_lib,
    family = binomial()
  )
  p_hat[test_idx] <- ps_fit$SL.predict
}

# (Optionally, examine the propensity score distribution)
summary(p_hat)
hist(p_hat, breaks = seq(0, 1, by=0.1), main = "Histogram of Propensity Scores")
axis(1, at = seq(0, 1, by=0.1))

# STEP 3: Construct the DR Pseudo-outcome
#   C = Y1_hat + (A / p_hat) * (Y - Y1_hat)
C <- Y1_hat + (A / p_hat) * (Y - Y1_hat)
summary(C)

# STEP 4: Fit the DR-Learner Model to Predict the Pseudo-outcome Using Only Reduced Covariates
dr_model <- SuperLearner(
  Y = C,
  X = X_reduced,    # Here, we use only age, inc, and educ as predictors
  newX = X_reduced,
  SL.library = sl_lib,
  family = gaussian()
)
Y1_tilde <- dr_model$SL.predict

# Check the distribution of the DR-learner predictions
hist(Y1_tilde, breaks = 40, col = "darkorange",
     main = "Histogram of DR-Learner Predictions (Counterfactual Y1)",
     xlab = "Predicted Counterfactual Outcome (Y1_tilde)")
summary(Y1_tilde)

# STEP 5: Evaluate the Counterfactual Prediction Error
# We use IPW weights for all subjects. 
weights_raw <- A / p_hat + (1 - A) / (1 - p_hat)
summary(weights_raw)
sd(weights_raw)
# IMPORTANT: Compute error using the pseudo-outcome C rather than the observed Y,
# because C represents our counterfactual target (Y1) through DR correction.
errors_raw <- weights_raw * (C - Y1_tilde)^2
error_est_raw <- mean(errors_raw)
se_raw <- sd(errors_raw) / sqrt(n)
ci_raw <- c(error_est_raw - 1.96 * se_raw, error_est_raw + 1.96 * se_raw)

# STEP 6: Output the Counterfactual Prediction Error Results
cat(sprintf("Counterfactual Prediction Error = %.2f\n95%% CI = [%.2f, %.2f]\n",
            error_est_raw, ci_raw[1], ci_raw[2]))

# Top 5 errors

# Step 1: Construct a data frame from the existing vectors.
df_errors <- data.frame(
  A = A,
  Y = Y,
  Y1_tilde = Y1_tilde,
  error = errors_raw
)

# Step 2: Compute the total error over all observations.
total_error <- sum(df_errors$error)

# Step 3: Identify the top 5 observations with the largest prediction error 
# and compute each observation's error contribution (in percentage).
top5_errors <- df_errors %>%
  arrange(desc(error)) %>%
  slice_head(n = 5) %>%
  mutate(Error_pct = (error / total_error) * 100)

# Step 4: Create a new data frame selecting only the relevant columns and format numbers.
df_display <- top5_errors %>%
  select(A, Y, Y1_tilde, error, Error_pct) %>%
  mutate(
    Y = ifelse(abs(as.numeric(Y)) >= 1e6,
               formatC(as.numeric(Y), format = "f", digits = 2, big.mark = ","),
               formatC(as.numeric(Y), format = "f", digits = 2)),
    Y1_tilde = ifelse(abs(as.numeric(Y1_tilde)) >= 1e6,
                      formatC(as.numeric(Y1_tilde), format = "f", digits = 2, big.mark = ","),
                      formatC(as.numeric(Y1_tilde), format = "f", digits = 2)),
    error = ifelse(abs(as.numeric(error)) >= 1e6,
                   formatC(as.numeric(error), format = "f", digits = 2, big.mark = ","),
                   formatC(as.numeric(error), format = "f", digits = 2)),
    Error_pct = formatC(Error_pct, format = "f", digits = 2)
  )

# Step 5: Create and display the flextable.
ft_top5 <- flextable(df_display) %>%
  set_header_labels(
    A = "Treatment (A)",
    Y = "Observed Outcome (Y)",
    Y1_tilde = "Predicted Counterfactual (Y1_tilde)",
    error = "Prediction Error",
    Error_pct = "Error Contribution (%)"
  ) %>%
  autofit() %>%
  set_caption("Top 5 Observations with Largest Prediction Error and Their Error Contribution")

print(ft_top5)
```

